---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mapAI: Positional Accuracy Improvement for Geospatial Data

<!-- badges: start -->
[![packageversion](https://img.shields.io/badge/Package%20version-1.0.0-orange.svg?style=flat-square)](https://github.com/kvantas/mapAI)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.15767080.svg)](https://doi.org/10.5281/zenodo.15767080)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![R-CMD-check](https://github.com/kvantas/mapAI/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/kvantas/mapAI/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The **mapAI** package delivers an open-source, mathematically rigorous framework for **Positional Accuracy Improvement (PAI)** of vector (`sf`) and raster (`terra`) geospatial data. Grounded in geodetic science, spatial statistics, and differential geometry, `mapAI` unifies classical conformal transformations, bivariate thin plate regression splines, spatial cross-validation schemes, and differential distortion diagnostics into a cohesive pair of spatial engines.

---

## Methodological Framework

Positional accuracy improvement addresses the systematic discrepancies between an approximate source spatial dataset (e.g., historical maps, digitized cadastral boundaries, unrectified aerial imagery) and actual reference coordinates established by geodetic ground control points (GCPs).

Given homologous point pairs with source coordinates $\mathbf{s} = (u, v)^T$ and true target coordinates $\mathbf{t} = (x, y)^T$, the objective is to estimate the vector displacement field:

$$\mathbf{d}(\mathbf{s}) = \begin{pmatrix} d_x(u, v) \\ d_y(u, v) \end{pmatrix} = \mathbf{t} - \mathbf{s}$$

Once trained, the model predicts continuous displacement vectors across space, allowing arbitrary vector geometries $\mathcal{G}_{\text{source}}$ and continuous or categorical raster grids $\mathcal{R}_{\text{source}}$ to be transformed into geometrically consistent spatial alignments.

```
+-----------------------------------------------------------------------------------+
|                              mapAI Architecture                                   |
+-----------------------------------------------------------------------------------+
|                                                                                   |
|   1. Data Ingestion & Georeferencing                                              |
|      read_gcp()  --->  Homologous points with optional CRS (sf / gcp S3)          |
|      read_map()  --->  Vector simple features (sf) or SpatRaster (terra)          |
|                                                                                   |
|   2. Model Training & Spatial Validation (Integrated)                             |
|      train_pai_model(gcp, method = ..., cv = list(...))                           |
|      * Global:  "helmert" (OLS / TLS / Procrustes), "lm"                          |
|      * Smooth:  "tps", "gam_biv"                                                  |
|      * TIN:     "tin_linear", "tin_akima"                                         |
|      * Hybrid:  "hybrid_helmert_tin", "hybrid_affine_tin"                         |
|      * Custom ML plugins via method = list(...)                                   |
|      * CV: random, spatial (k-means), spatial_block, spatial_buffered, etc.       |
|                                                                                   |
|   3. Dual-Engine Spatial Rectification                                            |
|      Vector Engine: transform_map(model, vector_map, repair_topology = TRUE)      |
|      Raster Engine: apply_pai_raster(model, raster, ext = "auto", mesh_step = 10) |
|                                                                                   |
|   4. Differential Diagnostics & Inversion Detection                               |
|      analyze_distortion()  ---> Metric tensor, singular values (a, b)             |
|      det(J) <= 0           ---> Topological fold-over & singularity detection     |
|      indicatrices()        ---> Tissot deformation ellipses                       |
+-----------------------------------------------------------------------------------+
```

---

## Supported PAI Transformation Models

`mapAI` includes a versatile hierarchy of models spanning from rigid geodetic adjustments to non-parametric splines and custom machine learning algorithms:

### 1. Helmert 2D Similarity Transformation (`"helmert"`)
A 4-parameter conformal transformation preserving local shape and angles:

$$\begin{pmatrix} x \\ y \end{pmatrix} = \begin{pmatrix} t_x \\ t_y \end{pmatrix} + \begin{pmatrix} a & -b \\ b & a \end{pmatrix} \begin{pmatrix} u \\ v \end{pmatrix}$$

Parameters include scale factor $s = \sqrt{a^2 + b^2}$, rotation angle $\theta = \operatorname{atan2}(b, a)$, and translation vector $(t_x, t_y)^T$. `mapAI` implements three estimators:

- **Ordinary Least Squares (`method = "ols"`):** Assumes measurement errors reside solely in the target coordinates.
- **Total Least Squares (`method = "tls"`):** Applies the perturbation to the full augmented matrix $[\mathbf{A} \mid \mathbf{d}]$, so error in the *source* coordinates is accounted for as well. This reduces the attenuation (regression-dilution) bias that OLS incurs when the source coordinates are themselves measured with error (Golub & Van Loan, 2013).
- **Scaled orthogonal Procrustes (`method = "procrustes"`):** For this 4-parameter conformal model, Procrustes minimises the same objective over the same parameter set as OLS and therefore returns the same estimate to machine precision. It is offered for comparison, not as an errors-in-variables method.

### 2. First-Order Polynomial Affine (`"lm"`)
Captures differential scale, uniform rotation, and linear affine shear along non-orthogonal axes:

$$d_x(u, v) = \beta_{0x} + \beta_{1x} u + \beta_{2x} v, \quad d_y(u, v) = \beta_{0y} + \beta_{1y} u + \beta_{2y} v$$

### 3. Thin Plate Splines (`"tps"`)
Minimizes residual sum of squares subject to a physical bending energy penalty (Wahba, 1990):

$$\mathcal{J}(f) = \sum_{i=1}^n \left(y_i - f(\mathbf{s}_i)\right)^2 + \lambda \iint_{\mathbb{R}^2} \left[ \left(\frac{\partial^2 f}{\partial u^2}\right)^2 + 2\left(\frac{\partial^2 f}{\partial u \partial v}\right)^2 + \left(\frac{\partial^2 f}{\partial v^2}\right)^2 \right] du\,dv$$

### 4. Bivariate GAM with Adaptive Spline Dimension (`"gam_biv"`)
A bivariate thin plate regression spline fitted via a multivariate Gaussian family (Wood, 2017):

$$\begin{pmatrix} d_x \\ d_y \end{pmatrix} \sim \mathcal{N}_2\left(\begin{pmatrix} \mu_x(u, v) \\ \mu_y(u, v) \end{pmatrix}, \boldsymbol{\Sigma}\right)$$

Smoothing parameters are selected by REML. `mapAI` incorporates an **adaptive basis dimension formula**:

$$k = \max\left(3, \min\left(k_{\max}, \lfloor 0.6 \cdot n_{\text{unique}} \rfloor\right)\right)$$

The lower bound prevents rank deficiency on small control point sets. The upper bound $k_{\max}$ (default 29) caps flexibility and binds for any network of 49 or more distinct points; if `mgcv::k.check()` shows the effective degrees of freedom approaching $k - 1$, pass a larger `k` or `k_max` to relax it.

### 5. Triangulated and Hybrid Models

- **`"tin_linear"`** — piecewise affine Delaunay triangulation with barycentric interpolation; exact at control points, with a global Helmert fallback outside the convex hull.
- **`"tin_akima"`** — smooth $C^1$ triangulated interpolation (Akima, 1996), avoiding the derivative kinks of `tin_linear`.
- **`"hybrid_helmert_tin"` / `"hybrid_affine_tin"`** — a global conformal or affine trend with triangulated residuals.

Note that *inside* the convex hull all three hybrid/TIN variants reduce to the same piecewise-affine map: for any affine base the base cancels against the barycentric weights. They differ only in how they extrapolate beyond the hull.

---

## Integrated Cross-Validation & Spatial Leakage Mitigation

Standard random $k$-fold cross-validation suffers from **optimism bias** when applied to spatial data, because spatial autocorrelation between nearby training and testing points leaks geographic information (Roberts et al., 2017). `mapAI` integrates rigorous cross-validation directly into `train_pai_model()` and exports `cv_pai_model()` / `assess_pai_model()`, supporting six distinct schemes:

1. **`"spatial_block"`:** Partitions space into regular rectangular geographic grid tiles, placing entire contiguous regions into independent folds.
2. **`"spatial_buffered"`:** Imposes an exclusion dead-zone buffer $d_{\text{buffer}}$ around validation blocks:

   $$\min_{\mathbf{s}_i \in \text{Test}} \|\mathbf{s}_j - \mathbf{s}_i\| > d_{\text{buffer}} \quad \forall \mathbf{s}_j \in \text{Train}$$

   Training points within the exclusion dead zone are pruned, which *reduces* autocorrelation leakage. It removes leakage only if $d_{\text{buffer}}$ exceeds the range of residual spatial autocorrelation; `mapAI` does not estimate that range, so the default is a geometric 5% of the map extent.
3. **`"spatial"`:** Spatial clustering of coordinates via $k$-means.
4. **`"random"`:** Standard random $k$-fold cross-validation.
5. **`"probability"`:** Single design-based train/test split.
6. **`"stratified"`:** Stratified cross-validation partitioned across displacement magnitude quantiles.

---

## Installation

Install the latest development version of `mapAI` from GitHub:

```r
# install.packages("pak")
pak::pak("kvantas/mapAI")
```

---

## Complete Hands-On Workflow

### 1. Data Ingestion and Visualizing GCPs

```{r data-creation, message=FALSE}
library(mapAI)
library(sf)
library(ggplot2)

# Generate a synthetic demonstration dataset featuring complex non-linear distortion
demo_data <- create_demo_data(type = "complex", seed = 42)

# Homologous points object
gcp_data <- demo_data$gcp
map_to_correct <- demo_data$map

# Inspect homologous points summary and plot displacement vectors
summary(gcp_data)
plot(gcp_data, title = "Ground Control Point Displacements")
```

### 2. Training with Integrated Spatial Block Cross-Validation

We train a bivariate GAM model while simultaneously assessing its out-of-sample performance using 5-fold regular spatial block cross-validation:

```{r train-model, message=FALSE}
# Train model with integrated spatial cross-validation
gam_model <- train_pai_model(
  gcp_data = gcp_data,
  method = "gam_biv",
  cv = list(validation_type = "spatial_block", k_folds = 5, seed = 42)
)

# Inspect model summary and cross-validation metrics
print(gam_model)
```

The headline out-of-sample metric is stored at `gam_model$cv_rmse_2d`, with the full assessment (including `Mean_RMSE_2D`, `SD_RMSE_2D` and the per-fold predictions) in `gam_model$cv`. It pools the squared residuals of every out-of-fold point, giving a more realistic estimate of predictive accuracy than in-sample residuals. Spatial schemes reduce, but do not eliminate, optimism from spatial autocorrelation.

### 3. Vector Map Transformation with Topology Validation

`transform_map()` extracts vertex coordinate matrices across all geometry types (`POINT`, `LINESTRING`, `POLYGON`, `MULTIPOINT`, `MULTILINESTRING`, `MULTIPOLYGON`), computes batch-vectorized predictions, reconstructs simple feature geometries, and automatically validates and repairs any self-intersections caused by non-linear warping (`repair_topology = TRUE`):

```{r transform-vector, message=FALSE, fig.show='hold'}
# Transform the distorted map
corrected_map <- transform_map(gam_model, map_to_correct, repair_topology = TRUE)

# Combine original and corrected features for comparison
map_to_correct$status <- "Original (Distorted)"
corrected_map$status <- "Corrected (PAI)"
comparison_data <- rbind(map_to_correct[, "status"], corrected_map[, "status"])

# Visualize overlay
ggplot(comparison_data) +
  geom_sf(aes(color = status, linetype = status), fill = NA, linewidth = 0.7) +
  scale_color_manual(name = "Map Status", values = c("Original (Distorted)" = "grey50", "Corrected (PAI)" = "#e41a1c")) +
  scale_linetype_manual(name = "Map Status", values = c("Original (Distorted)" = "dashed", "Corrected (PAI)" = "solid")) +
  labs(
    title = "Vector Positional Accuracy Improvement",
    subtitle = "Overlay of original distorted (dashed) and corrected (solid) features"
  ) +
  theme_minimal()
```

### 4. Pure In-Memory Raster Rectification

`apply_pai_raster()` brings PAI to raster imagery and scanned historical maps. Because raster warping requires **inverse (backward) mapping** to avoid raster gaps and collisions, `mapAI` implements a **damped fixed-point iterative coordinate inversion solver**:

$$\mathbf{s}^{(k+1)} = \mathbf{s}^{(k)} - \lambda \left(\mathbf{s}^{(k)} + \mathbf{d}\left(\mathbf{s}^{(k)}\right) - \mathbf{t}\right)$$

with damping factor $\lambda = 0.7$ and convergence tolerance $\epsilon = 10^{-4}$.

The interpolation method must match the data type: `"bilinear"` for continuous rasters (imagery, elevation models, scanned map tones), and `"near"` for discrete or categorical rasters, which relocates cells without altering their values.

```{r raster-rectification, message=FALSE, warning=FALSE, fig.show='hold'}
library(terra)

# Generate synthetic distorted raster matching the demonstration extent
demo_raster_data <- create_demo_data(type = "complex", seed = 42, raster = TRUE)
distorted_raster <- demo_raster_data$raster

# Rectify raster using nearest-neighbour resampling and automatic target extent
# calculation. The demo raster is a binary mask (0 = background, 1 = grid line),
# so an interpolating method such as "bilinear" would average neighbouring
# classes and introduce intermediate values absent from the source. "near"
# relocates cells without altering their values.
corrected_raster <- apply_pai_raster(
  gam_model,
  distorted_raster,
  method = "near",
  ext = "auto"
)

# Visualize original vs. rectified raster side-by-side on an identical scale
par(mfrow = c(1, 2))
terra::plot(distorted_raster, main = "Original (Distorted) Raster",
            col = c("grey90", "steelblue4"), range = c(0, 1))
terra::plot(corrected_raster, main = "Rectified Raster (mapAI)",
            col = c("grey90", "steelblue4"), range = c(0, 1))
par(mfrow = c(1, 1))
```

---

## Differential Distortion Analysis & Topological Inversion Detection

A key contribution of `mapAI` is its ability to diagnose and quantify local geometric deformations using differential geometry and Tissot's indicatrix theory (Tissot, 1881; Snyder, 1987).

Given numerical partial derivatives of the forward coordinate transformation:

$$\mathbf{J} = \begin{pmatrix} \frac{\partial f_x}{\partial x} & \frac{\partial f_x}{\partial y} \\ \frac{\partial f_y}{\partial x} & \frac{\partial f_y}{\partial y} \end{pmatrix}$$

`analyze_distortion()` evaluates:
- **Principal semi-axes ($a, b$):** The maximum and minimum local linear scale distortions (singular values of $\mathbf{J}$).
- **Signed Jacobian Determinant ($\det(\mathbf{J})$):**

  $$\det(\mathbf{J}) = \frac{\partial f_x}{\partial x}\frac{\partial f_y}{\partial y} - \frac{\partial f_x}{\partial y}\frac{\partial f_y}{\partial x}$$

  When $\det(\mathbf{J}) \le 0$, the mapping experiences **topological fold-over** or singular collapse. `mapAI` flags these conditions automatically via `is_inverted` and issues diagnostic warnings.
- **Maximum Angular Distortion:** $2 \arcsin\left(\frac{a - b}{a + b}\right)$
- **Airy-Kavrayskiy Criterion:** $\sqrt{\frac{1}{2}\left(\left(\ln \frac{a}{s_{\text{ref}}}\right)^2 + \left(\ln \frac{b}{s_{\text{ref}}}\right)^2\right)}$, where $s_{\text{ref}}$ is the reference linear scale (`reference_scale`)

```{r distortion-analysis, message=FALSE, warning=FALSE}
# Evaluate distortion metrics across the training points
distortion_field <- analyze_distortion(gam_model, gcp_data)
summary(distortion_field)

# Plot continuous areal distortion surface
plot(distortion_field, metric = "area_scale", diverging = TRUE) +
  labs(title = "Local Areal Distortion Scale")

# Plot Tissot's indicatrices showing local deformation ellipses
indicatrices(distortion_field)
```

---

## References

- Pebesma, E. (2018). Simple Features for R: Standardized Support for Spatial Vector Data. *The R Journal*, 10(1), 439-446. https://doi.org/10.32614/RJ-2018-009
- Roberts, D. R., Bahn, V., Ciuti, S., Boyce, M. S., Elith, J., Guillera-Arroita, G., Hauenstein, S., Lahoz-Monfort, J. J., Schröder, B., Thuiller, W., Warton, D. I., Wintle, B. A., Hartig, F., & Dormann, C. F. (2017). Cross-validation strategies for data with spatial, temporal, hierarchical, or phylogenetic structure. *Ecography*, 40(8), 913-929. https://doi.org/10.1111/ecog.02881
- Snyder, J. P. (1987). *Map Projections: A Working Manual*. U.S. Geological Survey Professional Paper 1395.
- Tissot, A. (1881). *Mémoire sur la représentation des surfaces et les projections des cartes géographiques*. Gauthier-Villars.
- Valavi, R., Elith, J., Lahoz-Monfort, J. J., & Guillera-Arroita, G. (2019). blockCV: An R package for generating spatially or environmentally separated folds for k-fold cross-validation of species distribution models. *Methods in Ecology and Evolution*, 10(2), 225-232. https://doi.org/10.1111/2041-210X.13107
- Vantas, K., & Mirkopoulou, E. (2025). *mapAI: An R Package for Positional Accuracy Improvement of Vector Maps*.
- Wahba, G. (1990). *Spline Models for Observational Data*. Society for Industrial and Applied Mathematics.
- Wolberg, G. (1990). *Digital Image Warping*. IEEE Computer Society Press.
- Wolf, P. R., & Ghilani, C. D. (2006). *Adjustment Computations: Spatial Data Analysis* (4th ed.). John Wiley & Sons.
- Wood, S. N. (2017). *Generalized Additive Models: An Introduction with R* (2nd ed.). Chapman & Hall/CRC.

---

## Documentation & Vignettes

Comprehensive guides and articles are available:
* [A Rigorous Methodological and Computational Framework for Positional Accuracy Improvement](docs/articles/mapAI_capabilities.md): In-depth treatise detailing theoretical foundations, mathematical formulations, spatial cross-validation schemes, vector/raster transformation engines, and differential distortion diagnostics.
* [Advanced Analysis with Basel and Frickthal Data](docs/articles/swiss_data.md): Case study on historical map georeferencing and residual error modeling using W. Haas's 1798 map of the Basel and Frickthal region.
* [Custom Model Training with mapAI](docs/articles/custom_model_training.md): Tutorial on integrating custom machine learning regression models (SVR, Random Forest, Neural Networks) into `mapAI`.

---

## Meta

* Bug reports, suggestions, and code are welcome via GitHub Issues.
* License: MIT.

