---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mapAI

<!-- badges: start -->
[![packageversion](https://img.shields.io/badge/Package%20version-1.0.0-orange.svg?style=flat-square)](https://github.com/kvantas/mapAI)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.15767080.svg)](https://doi.org/10.5281/zenodo.15767080)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![R-CMD-check](https://github.com/kvantas/mapAI/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/kvantas/mapAI/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The `mapAI` package is designed to provide a comprehensive and accessible PAI (positional accuracy improvement) framework for vector and raster geospatial data.

## Overview

The `mapAI` package provides a comprehensive and accessible framework for Positional Accuracy Improvement (PAI) of both vector (`sf`) and raster (`terra::SpatRaster`) geospatial data. This package’s main contributions are: 

(a) the unification of a set of PAI methods from classical adjustments to statistical and machine learning algorithms, within a framework engineered to modify the geometry of vector features and spatial alignment of raster maps; 

(b) the application of modern best practices regarding predictive accuracy assessment using various validation methods, and;

(c) the integration of distortion analysis into the PAI workflow for both vector and raster data, providing powerful diagnostics.

## Installation

You can install the development version of `mapAI` from [GitHub](https://github.com/) using the `pak` package:

``` r
# install.packages("pak")
pak::pak("kvantas/mapAI")
```

## Core Workflow: A Complete Example

This example demonstrates the primary workflow. We will first generate a synthetic dataset representing a distorted map and then use the package's functions to correct it.

### 1. Load Libraries and Create Demo Data

We begin by using `create_demo_data()` to generate a test case with complex, noisy distortions.

```{r data-creation}
library(mapAI)
library(sf)
library(ggplot2)

# Generate a shapefile and a GCPs CSV with complex noisy distortions
# The function returns a list containing the paths to these new files.
demo_data <- create_demo_data(type = "complex", seed = 1)


# plot the distortion data
plot(demo_data$gcp, main = "Homologous Points (GCPs)")

```

### 2. Read Data and Train a Model

We load the generated files and train a **Generalized Additive Model (`gam`)**, which is ideal for capturing the smooth, non-linear distortions present in the demo data.

```{r load-and-train}
# Load the homologous points (GCPs) and the distorted vector map
gcp_data <- demo_data$gcp
map_to_correct <- demo_data$map

# Train a bivariate GAM model using the GCPs
gam_model <- train_pai_model(gcp_data, "gam_biv")
```

### 3. Apply Correction and Visualize

We apply the trained model to our distorted grid. The resulting plot, which overlays the corrected grid on the original, provides a clear visual confirmation of what the model does to the distorted map.

```{r apply-and-visualize, fig.show='hold'}
# Apply the model to the distorted map
corrected_map <- transform_map(gam_model, map_to_correct)

# For easy plotting, add a 'status' column and combine the maps
map_to_correct$status <- "Original (Distorted)"
corrected_map$status <- "Corrected"
comparison_data <- rbind(map_to_correct[, "status"], corrected_map[, "status"])

# Create the final comparison plot
ggplot(comparison_data) +
  geom_sf(aes(color = status, linetype = status), fill = NA, linewidth = 0.7) +
  scale_color_manual(name = "Map Status", values = c("Original (Distorted)" = "grey50", "Corrected" = "#e41a1c")) +
  scale_linetype_manual(name = "Map Status", values = c("Original (Distorted)" = "dashed", "Corrected" = "solid")) +
  labs(title = "Positional Correction of a Distorted Grid",
       subtitle = "Overlay of original (dashed) and mapAI-corrected (solid) geometries") +
  theme_minimal()
```

### 4. Correcting Raster Maps (In-Memory)

The same trained PAI model can also be applied directly to rectify continuous or categorical raster data (e.g., scanned historical paper maps, aerial photographs, or digital elevation models) using `terra::SpatRaster`.

All raster transformations operate strictly in memory without creating temporary files:

```{r raster-correction, message=FALSE, warning=FALSE, fig.show='hold'}
library(terra)

# Generate demo data including an in-memory raster map
demo_raster_data <- create_demo_data(type = "complex", seed = 1, raster = TRUE)
distorted_raster <- demo_raster_data$raster

# Apply the trained model to correct the raster
# Uses inverse mapping with bilinear interpolation
corrected_raster <- transform_map(
  gam_model,
  distorted_raster,
  method = "bilinear"
)

# Visualize original vs. corrected raster side-by-side
par(mfrow = c(1, 2))
terra::plot(distorted_raster, main = "Original (Distorted) Raster", col = terrain.colors(50))
terra::plot(corrected_raster, main = "mapAI-Corrected Raster", col = terrain.colors(50))
par(mfrow = c(1, 1))
```

Key raster capabilities:
- **Inverse (Backward) Mapping**: Projects regular target grid cell centers backward into the source image space, eliminating holes and pixel collisions.
- **Continuous & Categorical Support**: Uses `"bilinear"` interpolation for continuous surfaces and `"near"` (nearest neighbor) for discrete thematic layers.
- **Fast Mesh Subsampling (`mesh_step`)**: Enables high-performance warping on large rasters with sub-pixel precision.
- **Pure In-Memory Operations**: Entirely preserves RAM efficiency with zero disk overhead.

---

## From Correction to Explanation: Advanced Distortion Analysis

A key challenge with data-driven models is understanding *what* they have learned. `mapAI` directly addresses this by providing tools to "open the black box" and analyze the properties of the learned transformation.

### 5. Quantify and Visualize the Distortion Field

The `analyze_distortion()` function computes local distortion metrics across the map space. This allows us to move from a simple visual assessment to a quantitative map of the distortions that the model learns.

```{r advanced-analysis, message=FALSE, warning=FALSE}
# 1. Analyze the distortion using our trained GAM model on GCPs
distortion_results <- analyze_distortion(gam_model, gcp_data)

# 2. Plot the distortion surfaces interpolating the values on gcp's location
plot(distortion_results,
     metric = "area_scale",
     diverging = TRUE) +
  labs(title = "Areal Distortion")

plot(
  distortion_results, metric = "max_shear"
) + 
  labs(title = "Maximum Shear Distortion (°)")
```

`analyze_distortion()` also evaluates distortion directly on in-memory `terra::SpatRaster` objects, returning an 8-layer raster containing all Tissot indicatrix metrics:

```{r raster-distortion, message=FALSE, warning=FALSE}
# Evaluate distortion across all cell centers of the raster
raster_distortion <- analyze_distortion(gam_model, newdata = distorted_raster)

# Plot key distortion metric layers directly from the raster
terra::plot(raster_distortion[[c("area_scale", "max_shear")]],
            main = c("Raster Areal Distortion", "Raster Max Shear (°)"))
```

## Meta

* Bug reports, suggestions, and code are welcome.

* License:
    + All code is licensed MIT.
