---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mapAI: Positional Accuracy Improvement for Geospatial Data

<!-- badges: start -->
[![packageversion](https://img.shields.io/badge/Package%20version-1.0.0-orange.svg?style=flat-square)](https://github.com/kvantas/mapAI)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.15767080.svg)](https://doi.org/10.5281/zenodo.15767080)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![R-CMD-check](https://github.com/kvantas/mapAI/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/kvantas/mapAI/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The **mapAI** package delivers an open-source, mathematically rigorous framework for **Positional Accuracy Improvement (PAI)** of vector (`sf`) and raster (`terra`) geospatial data. Grounded in geodetic science, spatial statistics, and differential geometry, `mapAI` unifies classical conformal transformations, bivariate thin plate regression splines, spatial cross-validation schemes, and differential distortion diagnostics into a cohesive pair of spatial engines.

---

## Methodological Framework

Positional accuracy improvement addresses the systematic discrepancies between an approximate source spatial dataset (e.g., historical maps, digitized cadastral boundaries, unrectified aerial imagery) and actual reference coordinates established by geodetic ground control points (GCPs).

Given homologous point pairs with source coordinates $\mathbf{s} = (u, v)^T$ and true target coordinates $\mathbf{t} = (x, y)^T$, the objective is to estimate the vector displacement field:

$$\mathbf{d}(\mathbf{s}) = \begin{pmatrix} d_x(u, v) \\ d_y(u, v) \end{pmatrix} = \mathbf{t} - \mathbf{s}$$

Once trained, the model predicts continuous displacement vectors across space, allowing arbitrary vector geometries $\mathcal{G}_{\text{source}}$ and continuous or categorical raster grids $\mathcal{R}_{\text{source}}$ to be transformed into geometrically consistent spatial alignments.

```
+-----------------------------------------------------------------------------------+
|                              mapAI Architecture                                   |
+-----------------------------------------------------------------------------------+
|                                                                                   |
|   1. Data Ingestion & Georeferencing                                              |
|      read_gcp()  --->  Homologous points with optional CRS (sf / gcp S3)          |
|      read_map()  --->  Vector simple features (sf) or SpatRaster (terra)          |
|                                                                                   |
|   2. Model Training & Spatial Validation (Integrated)                             |
|      train_pai_model(gcp, method = ..., cv = list(...))                           |
|      * Methods: "helmert" (OLS / TLS), "lm", "tps", "gam_biv", Custom ML          |
|      * CV: random, spatial (k-means), spatial_block, spatial_buffered, etc.       |
|                                                                                   |
|   3. Dual-Engine Spatial Rectification                                            |
|      Vector Engine: transform_map(model, vector_map, repair_topology = TRUE)      |
|      Raster Engine: apply_pai_raster(model, raster, ext = "auto", mesh_step = 10) |
|                                                                                   |
|   4. Differential Diagnostics & Inversion Detection                               |
|      analyze_distortion()  ---> Metric tensor, singular values (a, b)             |
|      det(J) <= 0           ---> Topological fold-over & singularity detection     |
|      indicatrices()        ---> Tissot deformation ellipses                       |
+-----------------------------------------------------------------------------------+
```

---

## Supported PAI Transformation Models

`mapAI` includes a versatile hierarchy of models spanning from rigid geodetic adjustments to non-parametric splines and custom machine learning algorithms:

### 1. Helmert 2D Similarity Transformation (`"helmert"`)
A 4-parameter conformal transformation preserving local shape and angles:

$$\begin{pmatrix} x \\ y \end{pmatrix} = \begin{pmatrix} t_x \\ t_y \end{pmatrix} + \begin{pmatrix} a & -b \\ b & a \end{pmatrix} \begin{pmatrix} u \\ v \end{pmatrix}$$

Parameters include scale factor $s = \sqrt{a^2 + b^2}$, rotation angle $\theta = \operatorname{atan2}(b, a)$, and translation vector $(t_x, t_y)^T$. `mapAI` implements both:
- **Ordinary Least Squares (`method = "ols"`):** Assumes measurement errors reside solely in target coordinates.
- **Total Least Squares / SVD Procrustes (`method = "tls"`):** Minimizes orthogonal Euclidean errors in both source and target coordinates simultaneously, eliminating attenuation bias (regression dilution) (Wolf & Ghilani, 2006).

### 2. First-Order Polynomial Affine (`"lm"`)
Captures differential scale, uniform rotation, and linear affine shear along non-orthogonal axes:

$$d_x(u, v) = \beta_{0x} + \beta_{1x} u + \beta_{2x} v, \quad d_y(u, v) = \beta_{0y} + \beta_{1y} u + \beta_{2y} v$$

### 3. Thin Plate Splines (`"tps"`)
Minimizes residual sum of squares subject to a physical bending energy penalty (Wahba, 1990):

$$\mathcal{J}(f) = \sum_{i=1}^n \left(y_i - f(\mathbf{s}_i)\right)^2 + \lambda \iint_{\mathbb{R}^2} \left[ \left(\frac{\partial^2 f}{\partial u^2}\right)^2 + 2\left(\frac{\partial^2 f}{\partial u \partial v}\right)^2 + \left(\frac{\partial^2 f}{\partial v^2}\right)^2 \right] du\,dv$$

### 4. Bivariate GAM with Adaptive Spline Dimension (`"gam_biv"`)
A bivariate thin plate regression spline fitted via a multivariate Gaussian family (Wood, 2017):

$$\begin{pmatrix} d_x \\ d_y \end{pmatrix} \sim \mathcal{N}_2\left(\begin{pmatrix} \mu_x(u, v) \\ \mu_y(u, v) \end{pmatrix}, \boldsymbol{\Sigma}\right)$$

`mapAI` incorporates an **adaptive basis dimension formula**:

$$k = \max\left(3, \min\left(29, \lfloor 0.6 \cdot n_{\text{unique}} \rfloor\right)\right)$$

which prevents rank deficiency when training with small sample sizes while maximizing flexibility on dense control point networks.

---

## Integrated Cross-Validation & Spatial Leakage Mitigation

Standard random $k$-fold cross-validation suffers from **optimism bias** when applied to spatial data, because spatial autocorrelation between nearby training and testing points leaks geographic information (Roberts et al., 2017). `mapAI` integrates rigorous cross-validation directly into `train_pai_model()` and exports `cv_pai_model()` / `assess_pai_model()`, supporting six distinct schemes:

1. **`"spatial_block"`:** Partitions space into regular rectangular geographic grid tiles, placing entire contiguous regions into independent folds.
2. **`"spatial_buffered"`:** Imposes an exclusion dead-zone buffer $d_{\text{buffer}}$ around validation blocks:

   $$\min_{\mathbf{s}_i \in \text{Test}} \|\mathbf{s}_j - \mathbf{s}_i\| > d_{\text{buffer}} \quad \forall \mathbf{s}_j \in \text{Train}$$

   Training points within the exclusion dead zone are pruned, eliminating autocorrelation leakage.
3. **`"spatial"`:** Spatial clustering of coordinates via $k$-means.
4. **`"random"`:** Standard random $k$-fold cross-validation.
5. **`"probability"`:** Single design-based train/test split.
6. **`"stratified"`:** Stratified cross-validation partitioned across displacement magnitude quantiles.

---

## Installation

Install the latest development version of `mapAI` from GitHub:

```r
# install.packages("pak")
pak::pak("kvantas/mapAI")
```

---

## Complete Hands-On Workflow

### 1. Data Ingestion and Visualizing GCPs

```{r data-creation, message=FALSE}
library(mapAI)
library(sf)
library(ggplot2)

# Generate a synthetic demonstration dataset featuring complex non-linear distortion
demo_data <- create_demo_data(type = "complex", seed = 42)

# Homologous points object
gcp_data <- demo_data$gcp
map_to_correct <- demo_data$map

# Inspect homologous points summary and plot displacement vectors
summary(gcp_data)
plot(gcp_data, title = "Ground Control Point Displacements")
```

### 2. Training with Integrated Spatial Block Cross-Validation

We train a bivariate GAM model while simultaneously assessing its out-of-sample performance using 5-fold regular spatial block cross-validation:

```{r train-model, message=FALSE}
# Train model with integrated spatial cross-validation
gam_model <- train_pai_model(
  gcp_data = gcp_data,
  method = "gam_biv",
  cv = list(validation_type = "spatial_block", k_folds = 5, seed = 42)
)

# Inspect model summary and cross-validation metrics
print(gam_model)
```

The out-of-sample cross-validation results (`cv_rmse_2d`) are stored directly in `gam_model$cv$summary`, providing a realistic estimate of predictive accuracy without data leakage.

### 3. Vector Map Transformation with Topology Validation

`transform_map()` extracts vertex coordinate matrices across all geometry types (`POINT`, `LINESTRING`, `POLYGON`, `MULTIPOINT`, `MULTILINESTRING`, `MULTIPOLYGON`), computes batch-vectorized predictions, reconstructs simple feature geometries, and automatically validates and repairs any self-intersections caused by non-linear warping (`repair_topology = TRUE`):

```{r transform-vector, message=FALSE, fig.show='hold'}
# Transform the distorted map
corrected_map <- transform_map(gam_model, map_to_correct, repair_topology = TRUE)

# Combine original and corrected features for comparison
map_to_correct$status <- "Original (Distorted)"
corrected_map$status <- "Corrected (PAI)"
comparison_data <- rbind(map_to_correct[, "status"], corrected_map[, "status"])

# Visualize overlay
ggplot(comparison_data) +
  geom_sf(aes(color = status, linetype = status), fill = NA, linewidth = 0.7) +
  scale_color_manual(name = "Map Status", values = c("Original (Distorted)" = "grey50", "Corrected (PAI)" = "#e41a1c")) +
  scale_linetype_manual(name = "Map Status", values = c("Original (Distorted)" = "dashed", "Corrected (PAI)" = "solid")) +
  labs(
    title = "Vector Positional Accuracy Improvement",
    subtitle = "Overlay of original distorted (dashed) and corrected (solid) features"
  ) +
  theme_minimal()
```

### 4. Pure In-Memory Raster Rectification

`apply_pai_raster()` brings PAI to raster imagery and scanned historical maps. Because raster warping requires **inverse (backward) mapping** to avoid raster gaps and collisions, `mapAI` implements a **damped fixed-point iterative coordinate inversion solver**:

$$\mathbf{s}^{(k+1)} = \mathbf{s}^{(k)} - \lambda \left(\mathbf{s}^{(k)} + \mathbf{d}\left(\mathbf{s}^{(k)}\right) - \mathbf{t}\right)$$

with damping factor $\lambda = 0.7$ and convergence tolerance $\epsilon = 10^{-4}$.

```{r raster-rectification, message=FALSE, warning=FALSE, fig.show='hold'}
library(terra)

# Generate synthetic distorted raster matching the demonstration extent
demo_raster_data <- create_demo_data(type = "complex", seed = 42, raster = TRUE)
distorted_raster <- demo_raster_data$raster

# Rectify raster using bilinear interpolation and automatic target extent calculation
corrected_raster <- apply_pai_raster(
  gam_model,
  distorted_raster,
  method = "bilinear",
  ext = "auto"
)

# Visualize original vs. rectified raster side-by-side
par(mfrow = c(1, 2))
terra::plot(distorted_raster, main = "Original (Distorted) Raster", col = terrain.colors(50))
terra::plot(corrected_raster, main = "Rectified Raster (mapAI)", col = terrain.colors(50))
par(mfrow = c(1, 1))
```

---

## Differential Distortion Analysis & Topological Inversion Detection

A key contribution of `mapAI` is its ability to diagnose and quantify local geometric deformations using differential geometry and Tissot's indicatrix theory (Tissot, 1881; Snyder, 1987).

Given numerical partial derivatives of the forward coordinate transformation:

$$\mathbf{J} = \begin{pmatrix} \frac{\partial f_x}{\partial x} & \frac{\partial f_x}{\partial y} \\ \frac{\partial f_y}{\partial x} & \frac{\partial f_y}{\partial y} \end{pmatrix}$$

`analyze_distortion()` evaluates:
- **Principal semi-axes ($a, b$):** The maximum and minimum local linear scale distortions (singular values of $\mathbf{J}$).
- **Signed Jacobian Determinant ($\det(\mathbf{J})$):**

  $$\det(\mathbf{J}) = \frac{\partial f_x}{\partial x}\frac{\partial f_y}{\partial y} - \frac{\partial f_x}{\partial y}\frac{\partial f_y}{\partial x}$$

  When $\det(\mathbf{J}) \le 0$, the mapping experiences **topological fold-over** or singular collapse. `mapAI` flags these conditions automatically via `is_inverted` and issues diagnostic warnings.
- **Maximum Angular Distortion:** $2 \arcsin\left(\frac{a - b}{a + b}\right)$
- **Airy-Kavrayskiy Criterion:** $\frac{1}{2}\left((\ln a)^2 + (\ln b)^2\right)$

```{r distortion-analysis, message=FALSE, warning=FALSE}
# Evaluate distortion metrics across the training points
distortion_field <- analyze_distortion(gam_model, gcp_data)
summary(distortion_field)

# Plot continuous areal distortion surface
plot(distortion_field, metric = "area_scale", diverging = TRUE) +
  labs(title = "Local Areal Distortion Scale")

# Plot Tissot's indicatrices showing local deformation ellipses
indicatrices(distortion_field)
```

---

## References

- Pebesma, E. (2018). Simple Features for R: Standardized Support for Spatial Vector Data. *The R Journal*, 10(1), 439-446. https://doi.org/10.32614/RJ-2018-009
- Roberts, D. R., Bahn, V., Ciuti, S., Boyce, M. S., Elith, J., Guillera-Arroita, G., Hauenstein, S., Lahoz-Monfort, J. J., Schröder, B., Thuiller, W., Warton, D. I., Wintle, B. A., Hartig, F., & Dormann, C. F. (2017). Cross-validation strategies for data with spatial, temporal, or phylogenetic dependence. *Ecography*, 40(8), 913-929. https://doi.org/10.1111/ecog.02881
- Snyder, J. P. (1987). *Map Projections: A Working Manual*. U.S. Geological Survey Professional Paper 1395.
- Tissot, A. (1881). *Mémoire sur la représentation des surfaces et les projections des cartes géographiques*. Gauthier-Villars.
- Valavi, R., Elith, J., Lahoz-Monfort, J. J., & Guillera-Arroita, G. (2019). blockCV: An R package for generating spatially or environmentally separated folds for k-fold cross-validation of species distribution models. *Methods in Ecology and Evolution*, 10(2), 225-232. https://doi.org/10.1111/2041-210X.13107
- Vantas, K., & Mirkopoulou, E. (2025). *mapAI: An R Package for Positional Accuracy Improvement of Vector Maps*.
- Wahba, G. (1990). *Spline Models for Observational Data*. Society for Industrial and Applied Mathematics.
- Wolberg, G. (1990). *Digital Image Warping*. IEEE Computer Society Press.
- Wolf, P. R., & Ghilani, C. D. (2006). *Adjustment Computations: Spatial Data Analysis* (4th ed.). John Wiley & Sons.
- Wood, S. N. (2017). *Generalized Additive Models: An Introduction with R* (2nd ed.). Chapman & Hall/CRC.

---

## Meta

* Bug reports, suggestions, and code are welcome via GitHub Issues.
* License: MIT.

