---
title: "mapAI Corner Case Behaviour Report"
subtitle: "Synthetic stress tests for the vector and raster engines"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: kable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
```

## What this document is

A behaviour probe, not a unit test suite. Every case below is deliberately
awkward — degenerate, empty, out-of-domain, or numerically nasty — and the
question is always the same: **does `mapAI` do the right thing, and does it say
so?**

The distinction that matters is between the three outcomes:

| Outcome | Meaning |
|---|---|
| **ok** | Completed with no error and no warning |
| **warning** | Completed, but told the user something was off |
| **error** | Refused to proceed |

For a corner case, `ok` is not automatically good. A function that quietly
returns a plausible-looking wrong answer is worse than one that errors. Each case
below therefore declares what it *expects*, and the summary flags any mismatch.

All data is synthetic and generated in this document; nothing is read from disk.

```{r libs}
library(mapAI)
library(sf)
library(terra)
```

## Harness

```{r harness}
cc <- new.env()
cc$rows <- list()

#' Run one corner case, capturing its outcome without aborting the report.
#'
#' @param case    short description of what is being probed
#' @param expr    the expression to evaluate
#' @param expect  "ok", "warning" or "error" -- the defensible behaviour
#' @param section grouping label
#' @param probe   optional function(value) returning a one-line string used to
#'                inspect the result beyond its mere existence
check <- function(case, expr, expect = c("ok", "warning", "error"),
                  section = "", probe = NULL) {
  expect <- match.arg(expect)
  warns <- character(0)
  err <- NULL

  val <- withCallingHandlers(
    tryCatch(expr, error = function(e) { err <<- conditionMessage(e); NULL }),
    warning = function(w) {
      warns <<- c(warns, conditionMessage(w))
      invokeRestart("muffleWarning")
    },
    message = function(m) invokeRestart("muffleMessage")
  )

  observed <- if (!is.null(err)) "error" else if (length(warns)) "warning" else "ok"

  detail <- if (!is.null(err)) {
    err
  } else if (length(warns)) {
    warns[1]
  } else if (!is.null(probe)) {
    tryCatch(probe(val), error = function(e) paste("probe failed:", conditionMessage(e)))
  } else {
    ""
  }

  # A warning where only "ok" was expected is still acceptable if it is
  # informative; the verdict flags the cases worth looking at.
  verdict <- if (observed == expect) {
    "as expected"
  } else if (expect == "ok" && observed == "warning") {
    "extra warning"
  } else if (expect %in% c("warning", "error") && observed == "ok") {
    "SILENT"
  } else {
    "MISMATCH"
  }

  cc$rows[[length(cc$rows) + 1L]] <- data.frame(
    section  = section,
    case     = case,
    expected = expect,
    observed = observed,
    verdict  = verdict,
    detail   = substr(gsub("\\s+", " ", detail), 1, 130),
    stringsAsFactors = FALSE
  )

  invisible(val)
}

report <- function(section) {
  df <- do.call(rbind, cc$rows)
  df <- df[df$section == section, c("case", "expected", "observed", "verdict", "detail")]
  rownames(df) <- NULL
  knitr::kable(df)
}
```

## Synthetic data generators

```{r generators}
# A regular GCP grid with a chosen smooth warp. Coordinates are deliberately
# large (Swiss-grid magnitude) for the numerical-precision cases.
make_gcp <- function(n = 8, span = 1000, origin = c(0, 0),
                     warp = c("smooth", "identity", "fold", "tiny"),
                     noise = 0) {
  warp <- match.arg(warp)
  g <- expand.grid(
    source_x = seq(origin[1], origin[1] + span, length.out = n),
    source_y = seq(origin[2], origin[2] + span, length.out = n)
  )
  u <- (g$source_x - origin[1]) / span
  v <- (g$source_y - origin[2]) / span

  d <- switch(warp,
    identity = list(dx = rep(0, nrow(g)), dy = rep(0, nrow(g))),
    smooth   = list(dx = span * (0.02 * u + 0.015 * u * v),
                    dy = span * (0.018 * v - 0.012 * u * v)),
    # Strong enough to reverse orientation over part of the domain.
    fold     = list(dx = -1.8 * span * u, dy = rep(0, nrow(g))),
    tiny     = list(dx = rep(1e-9, nrow(g)), dy = rep(1e-9, nrow(g)))
  )

  if (noise > 0) {
    d$dx <- d$dx + rnorm(nrow(g), 0, noise)
    d$dy <- d$dy + rnorm(nrow(g), 0, noise)
  }

  out <- data.frame(
    source_x = g$source_x,
    source_y = g$source_y,
    target_x = g$source_x + d$dx,
    target_y = g$source_y + d$dy,
    dx = d$dx,
    dy = d$dy
  )
  class(out) <- c("gcp", "data.frame")
  out
}

set.seed(1)
gcp_ok    <- make_gcp(8, warp = "smooth", noise = 0.5)
gcp_ident <- make_gcp(8, warp = "identity")
gcp_fold  <- make_gcp(8, warp = "fold")
gcp_big   <- make_gcp(8, span = 60000, origin = c(2600000, 1200000), warp = "smooth")

model_ok   <- train_pai_model(gcp_ok,   method = "lm")
model_gam  <- train_pai_model(gcp_ok,   method = "gam_biv")
model_tin  <- train_pai_model(gcp_ok,   method = "tin_linear")
model_ident <- train_pai_model(gcp_ident, method = "lm")

# Bounding box of the control network, used to place geometries inside/outside.
hull <- c(xmin = min(gcp_ok$source_x), xmax = max(gcp_ok$source_x),
          ymin = min(gcp_ok$source_y), ymax = max(gcp_ok$source_y))
hull
```

---

# 1. Model training

Degenerate control networks. The recurring question: does a network that cannot
support a model produce a clear refusal, or a fitted object that looks fine?

```{r sec-model}
sec <- "1. Model training"

# --- too few points for the model's parameter count ---
# Note rows 1:3 of the grid share a y value, so pick a genuine triangle.
gcp_1 <- gcp_ok[1, ]
gcp_2 <- gcp_ok[1:2, ]
gcp_3 <- gcp_ok[c(1, 2, 9), ]
class(gcp_3) <- c("gcp", "data.frame")

# Probe the fitted coefficients: a rank-deficient lm returns NA slopes, and a
# model with NA coefficients predicts NA. "Trained successfully" is not enough.
coef_probe <- function(m) {
  cf <- c(stats::coef(m$model$model_dx), stats::coef(m$model$model_dy))
  sprintf("%d of %d coefficients are NA", sum(is.na(cf)), length(cf))
}

check("1 GCP, lm (3 params per axis)", train_pai_model(gcp_1, "lm"),
      expect = "error", section = sec, probe = coef_probe)
check("2 GCPs, helmert (4 params)", train_pai_model(gcp_2, "helmert"),
      expect = "ok", section = sec,
      probe = function(m) sprintf("scale = %.4f, df = %d",
                                  m$model$parameters["scale"], m$model$df))
check("3 non-collinear GCPs, lm (exactly determined, 0 residual df)",
      train_pai_model(gcp_3, "lm"), expect = "ok", section = sec)
check("3 GCPs, gam_biv (below the basis dimension)",
      train_pai_model(gcp_3, "gam_biv"), expect = "error", section = sec)
check("3 non-collinear GCPs, tin_linear (1 triangle)",
      train_pai_model(gcp_3, "tin_linear"), expect = "ok", section = sec)

# --- collinear control points: no 2D information at all ---
gcp_line <- gcp_ok[gcp_ok$source_y == gcp_ok$source_y[1], ]
check(sprintf("collinear GCPs (n = %d, all same y), lm", nrow(gcp_line)),
      train_pai_model(gcp_line, "lm"), expect = "error", section = sec,
      probe = coef_probe)
check("collinear GCPs, helmert", train_pai_model(gcp_line, "helmert"),
      expect = "ok", section = sec)
check("collinear GCPs, tin_linear (no valid triangulation)",
      train_pai_model(gcp_line, "tin_linear"), expect = "error", section = sec)

# --- all control points at one location ---
gcp_dup <- gcp_ok[rep(1, 10), ]
check("10 identical GCPs, helmert", train_pai_model(gcp_dup, "helmert"),
      expect = "error", section = sec)
check("10 identical GCPs, lm", train_pai_model(gcp_dup, "lm"),
      expect = "error", section = sec, probe = coef_probe)

# --- duplicated coordinates mixed with valid ones ---
gcp_partdup <- rbind(gcp_ok, gcp_ok[1:5, ])
class(gcp_partdup) <- c("gcp", "data.frame")
check("GCPs with 5 exact duplicates, gam_biv",
      train_pai_model(gcp_partdup, "gam_biv"), expect = "ok", section = sec)
check("GCPs with 5 exact duplicates, tin_akima",
      train_pai_model(gcp_partdup, "tin_akima"), expect = "ok", section = sec)

# --- NA in the control network ---
gcp_na <- gcp_ok
gcp_na$dx[3] <- NA_real_
gcp_na$target_x[3] <- NA_real_
check("GCP with NA displacement, lm", train_pai_model(gcp_na, "lm"),
      expect = "error", section = sec,
      probe = function(m) sprintf("fitted on %d of %d GCPs (row silently dropped)",
                                  stats::nobs(m$model$model_dx), nrow(gcp_na)))
# assess_pai_model() rejects the same data outright, so the two entry points
# disagree about whether an NA control point is acceptable.
check("GCP with NA displacement, assess_pai_model",
      assess_pai_model(gcp_na, "lm", validation_type = "random", k_folds = 3),
      expect = "error", section = sec)

# --- zero displacement everywhere ---
# A map needing no correction is a legitimate input, and the response has zero
# variance. mgcv::mvn cannot estimate a residual covariance from it.
check("identity warp (all displacements 0), lm",
      train_pai_model(gcp_ident, "lm"), expect = "ok", section = sec)
check("identity warp (all displacements 0), helmert",
      train_pai_model(gcp_ident, "helmert"), expect = "ok", section = sec)
check("identity warp (all displacements 0), gam_biv",
      train_pai_model(gcp_ident, "gam_biv"), expect = "error", section = sec)
# A constant field has no residual variance for mgcv::mvn(d = 2) to model, so a
# clear refusal naming a method that CAN fit it is the right outcome.
check("near-identity warp (displacements 1e-9), gam_biv",
      train_pai_model(make_gcp(8, warp = "tiny"), "gam_biv"),
      expect = "error", section = sec)

# A map offset by a constant shift is an ordinary real-world case: the
# displacement is non-zero but has zero variance in both components.
gcp_shift <- gcp_ident
gcp_shift$dx <- 25; gcp_shift$dy <- -10
gcp_shift$target_x <- gcp_shift$source_x + 25
gcp_shift$target_y <- gcp_shift$source_y - 10
class(gcp_shift) <- c("gcp", "data.frame")

check("pure translation (constant dx/dy), helmert",
      train_pai_model(gcp_shift, "helmert"), expect = "ok", section = sec,
      probe = function(m) sprintf("tx = %.2f, ty = %.2f (true 25, -10)",
                                  m$model$parameters["tx"],
                                  m$model$parameters["ty"]))
check("pure translation (constant dx/dy), lm",
      train_pai_model(gcp_shift, "lm"), expect = "ok", section = sec)
check("pure translation (constant dx/dy), gam_biv",
      train_pai_model(gcp_shift, "gam_biv"), expect = "error", section = sec,
      probe = function(x) "refused, message names helmert/lm")
check("pure translation (constant dx/dy), tps",
      train_pai_model(gcp_shift, "tps"), expect = "ok", section = sec)

# --- orientation-reversing warp ---
check("fold-inducing warp, tin_linear (should flag inverted facets)",
      train_pai_model(gcp_fold, "tin_linear"), expect = "warning", section = sec)

# --- very large coordinates: numerical conditioning ---
check("Swiss-grid magnitude coordinates (~2.6e6), helmert",
      train_pai_model(gcp_big, "helmert"), expect = "ok", section = sec,
      probe = function(m) sprintf("scale = %.10f (true 1.0 + warp)",
                                  m$model$parameters["scale"]))

report(sec)
```

---

# 2. Vector transformation

## 2.1 Geometry type coverage

Every OGC simple-feature type the package claims to support, plus the empty and
degenerate forms of each.

```{r sec-vec-types}
sec <- "2.1 Geometry types"

crs_l <- 3857  # a projected CRS, so assert_projected_crs() is satisfied
inside <- function(fx, fy) {
  # map unit-square coordinates into the control network footprint
  c(hull["xmin"] + fx * (hull["xmax"] - hull["xmin"]),
    hull["ymin"] + fy * (hull["ymax"] - hull["ymin"]))
}

p1 <- inside(0.2, 0.2); p2 <- inside(0.8, 0.3); p3 <- inside(0.6, 0.8)

mk <- function(g) st_sf(id = seq_along(st_sfc(g)), geometry = st_sfc(g, crs = crs_l))

sf_point  <- mk(st_point(p1))
sf_mpoint <- mk(st_multipoint(rbind(p1, p2, p3)))
sf_line   <- mk(st_linestring(rbind(p1, p2, p3)))
sf_mline  <- mk(st_multilinestring(list(rbind(p1, p2), rbind(p2, p3))))

ring <- function(cx, cy, r) {
  th <- seq(0, 2 * pi, length.out = 33)
  m <- unname(cbind(cx + r * cos(th), cy + r * sin(th)))
  m[nrow(m), ] <- m[1, ]   # close the ring exactly; 2*pi != 0 in floating point
  m
}
ctr <- inside(0.5, 0.5); rad <- 0.2 * (hull["xmax"] - hull["xmin"])
sf_poly  <- mk(st_polygon(list(ring(ctr[1], ctr[2], rad))))
sf_hole  <- mk(st_polygon(list(ring(ctr[1], ctr[2], rad),
                               ring(ctr[1], ctr[2], rad / 3)[33:1, ])))
sf_mpoly <- mk(st_multipolygon(list(
  list(ring(p1[1], p1[2], rad / 2)),
  list(ring(p2[1], p2[2], rad / 2))
)))

for (nm in c("POINT", "MULTIPOINT", "LINESTRING", "MULTILINESTRING",
             "POLYGON", "POLYGON with hole", "MULTIPOLYGON")) {
  obj <- switch(nm,
    "POINT" = sf_point, "MULTIPOINT" = sf_mpoint, "LINESTRING" = sf_line,
    "MULTILINESTRING" = sf_mline, "POLYGON" = sf_poly,
    "POLYGON with hole" = sf_hole, "MULTIPOLYGON" = sf_mpoly)
  check(nm, transform_map(model_ok, obj), expect = "ok", section = sec,
        probe = function(x) sprintf("%d feature(s), valid = %s",
                                    nrow(x), all(st_is_valid(x))))
}

# Interior ring must survive as an interior ring, not be dropped or merged.
check("POLYGON with hole keeps its interior ring",
      transform_map(model_ok, sf_hole), expect = "ok", section = sec,
      probe = function(x) sprintf("n rings after = %d (input 2)",
                                  length(st_geometry(x)[[1]])))

# --- empty and degenerate geometries ---
sf_empty_pt   <- mk(st_point())
sf_empty_poly <- mk(st_polygon())
sf_zero_len   <- mk(st_linestring(rbind(p1, p1)))
sf_zero_area  <- mk(st_polygon(list(rbind(p1, p2, p1, p1))))

check("empty POINT (no finite coordinates)", transform_map(model_ok, sf_empty_pt),
      expect = "error", section = sec)
check("empty POLYGON", transform_map(model_ok, sf_empty_poly),
      expect = "ok", section = sec)
check("zero-length LINESTRING (two identical vertices)",
      transform_map(model_ok, sf_zero_len), expect = "ok", section = sec)
check("zero-area POLYGON (collapsed sliver)",
      transform_map(model_ok, sf_zero_area), expect = "ok", section = sec,
      probe = function(x) sprintf("area_new = %s",
                                  format(as.numeric(st_area(x)), digits = 3)))

# --- zero-row layer ---
check("sf object with 0 features",
      transform_map(model_ok, sf_poly[0, ]), expect = "ok", section = sec,
      probe = function(x) sprintf("nrow = %d", nrow(x)))

# --- mixed geometry types in one layer ---
sf_mixed <- rbind(
  st_sf(id = 1, geometry = st_sfc(st_point(p1), crs = crs_l)),
  st_sf(id = 2, geometry = st_sfc(st_polygon(list(ring(ctr[1], ctr[2], rad))),
                                  crs = crs_l))
)
check("mixed POINT + POLYGON in one layer (area_new must still appear)",
      transform_map(model_ok, sf_mixed), expect = "ok", section = sec,
      probe = function(x) sprintf("has area_new = %s", "area_new" %in% names(x)))

report(sec)
```

## 2.2 Topology

```{r sec-vec-topo}
sec <- "2.2 Topology"

# Already-invalid input: a bow-tie.
bow <- st_sf(id = 1, geometry = st_sfc(st_polygon(list(rbind(
  inside(0.3, 0.3), inside(0.7, 0.7), inside(0.7, 0.3),
  inside(0.3, 0.7), inside(0.3, 0.3)
))), crs = crs_l))

check("self-intersecting input, repair_topology = TRUE",
      transform_map(model_ok, bow, repair_topology = TRUE),
      expect = "ok", section = sec,
      probe = function(x) sprintf("valid after = %s", all(st_is_valid(x))))
check("self-intersecting input, repair_topology = FALSE",
      transform_map(model_ok, bow, repair_topology = FALSE),
      expect = "ok", section = sec,
      probe = function(x) sprintf("valid after = %s (repair off)",
                                  all(st_is_valid(x))))

# A warp strong enough to fold the domain: does the vector engine notice?
model_fold <- suppressWarnings(train_pai_model(gcp_fold, "lm"))
check("fold-inducing model applied to a polygon",
      transform_map(model_fold, sf_poly, repair_topology = TRUE),
      expect = "ok", section = sec,
      probe = function(x) {
        a <- as.numeric(st_area(x))
        sprintf("area %.1f -> %.1f, valid = %s",
                as.numeric(st_area(sf_poly)), a, all(st_is_valid(x)))
      })

report(sec)
```

## 2.3 Domain: inside, outside, and across the hull

The TIN family interpolates inside the convex hull and extrapolates outside it
with a different rule. A geometry that straddles the hull boundary crosses that
seam.

```{r sec-vec-domain}
sec <- "2.3 Domain and extrapolation"

far <- function(mult) {
  w <- hull["xmax"] - hull["xmin"]
  st_sf(id = 1, geometry = st_sfc(st_polygon(list(rbind(
    c(hull["xmax"] + mult * w,       hull["ymin"]),
    c(hull["xmax"] + (mult + 0.3) * w, hull["ymin"]),
    c(hull["xmax"] + (mult + 0.3) * w, hull["ymin"] + 0.3 * w),
    c(hull["xmax"] + mult * w,       hull["ymin"] + 0.3 * w),
    c(hull["xmax"] + mult * w,       hull["ymin"])
  ))), crs = crs_l))
}

disp_of <- function(x, src) {
  cs <- st_coordinates(src)[1, 1:2]; ct <- st_coordinates(x)[1, 1:2]
  sprintf("corner moved %.2f units", sqrt(sum((ct - cs)^2)))
}

check("polygon just outside the hull, gam_biv",
      transform_map(model_gam, far(0.1)), expect = "ok", section = sec,
      probe = function(x) disp_of(x, far(0.1)))
check("polygon 10x the hull width away, gam_biv (extrapolation)",
      transform_map(model_gam, far(10)), expect = "ok", section = sec,
      probe = function(x) disp_of(x, far(10)))
check("polygon 10x away, tin_linear (Helmert fallback)",
      transform_map(model_tin, far(10)), expect = "ok", section = sec,
      probe = function(x) disp_of(x, far(10)))

# A line crossing the hull boundary samples both sides of the TIN seam.
crossing <- st_sf(id = 1, geometry = st_sfc(st_linestring(rbind(
  inside(0.5, 0.5),
  c(hull["xmax"] + 0.5 * (hull["xmax"] - hull["xmin"]), inside(0.5, 0.5)[2])
)), crs = crs_l))

check("line crossing the TIN convex hull boundary",
      transform_map(model_tin, crossing), expect = "ok", section = sec,
      probe = function(x) {
        d <- st_coordinates(x)[, 1:2] - st_coordinates(crossing)[, 1:2]
        sprintf("endpoint displacements: %.2f and %.2f",
                sqrt(sum(d[1, ]^2)), sqrt(sum(d[nrow(d), ]^2)))
      })

report(sec)
```

## 2.4 CRS handling

```{r sec-vec-crs}
sec <- "2.4 CRS"

sf_nocrs <- st_sf(id = 1, geometry = st_sfc(st_polygon(list(
  ring(ctr[1], ctr[2], rad)))))
check("map with no CRS", transform_map(model_ok, sf_nocrs),
      expect = "warning", section = sec)

# Geographic coordinates: the model is planar Cartesian, so degrees are wrong.
sf_lonlat <- st_sf(id = 1, geometry = st_sfc(st_polygon(list(
  rbind(c(8.0, 47.0), c(8.1, 47.0), c(8.1, 47.1), c(8.0, 47.1), c(8.0, 47.0))
)), crs = 4326))
check("map in geographic (lon/lat) coordinates",
      transform_map(model_ok, sf_lonlat), expect = "warning", section = sec)

# AOI in a different CRS: must be reprojected, and the reprojection must be used.
aoi_other <- st_transform(
  st_sf(id = 1, geometry = st_sfc(st_polygon(list(
    ring(ctr[1], ctr[2], rad * 3))), crs = crs_l)), 4326)
check("AOI supplied in a different CRS than the map",
      transform_map(model_ok, sf_poly, aoi = aoi_other),
      expect = "ok", section = sec,
      probe = function(x) sprintf("%d feature(s) returned", nrow(x)))

# AOI that misses the map entirely.
aoi_miss <- st_sf(id = 1, geometry = st_sfc(st_polygon(list(
  ring(hull["xmax"] + 50 * rad, hull["ymax"] + 50 * rad, rad))), crs = crs_l))
check("AOI disjoint from the map",
      transform_map(model_ok, sf_poly, aoi = aoi_miss),
      expect = "ok", section = sec,
      probe = function(x) sprintf("%d feature(s) returned", nrow(x)))

report(sec)
```

---

# 3. Raster rectification

## 3.1 Shape and value degeneracies

```{r sec-rast-shape}
sec <- "3.1 Raster shape and values"

base_ext <- ext(hull["xmin"], hull["xmax"], hull["ymin"], hull["ymax"])

mk_rast <- function(nr, nc, vals = NULL, e = base_ext, crs_in = crs_l) {
  r <- rast(e, nrows = nr, ncols = nc)
  if (!is.null(crs_in)) terra::crs(r) <- paste0("EPSG:", crs_in)
  values(r) <- if (is.null(vals)) seq_len(ncell(r)) else vals
  r
}

check("1 x 1 raster", apply_pai_raster(model_ok, mk_rast(1, 1)),
      expect = "ok", section = sec,
      probe = function(r) sprintf("dim %s", paste(dim(r), collapse = "x")))
check("1 x 50 raster (single row)", apply_pai_raster(model_ok, mk_rast(1, 50)),
      expect = "ok", section = sec,
      probe = function(r) sprintf("dim %s", paste(dim(r), collapse = "x")))
check("50 x 1 raster (single column)", apply_pai_raster(model_ok, mk_rast(50, 1)),
      expect = "ok", section = sec,
      probe = function(r) sprintf("dim %s", paste(dim(r), collapse = "x")))

r_allna <- mk_rast(20, 20); values(r_allna) <- NA_real_
check("all-NA raster", apply_pai_raster(model_ok, r_allna),
      expect = "ok", section = sec,
      probe = function(r) sprintf("non-NA out = %d", sum(!is.na(values(r)[, 1]))))

r_const <- mk_rast(20, 20, vals = rep(7, 400))
check("constant-valued raster", apply_pai_raster(model_ok, r_const),
      expect = "ok", section = sec,
      probe = function(r) {
        v <- values(r)[, 1]
        sprintf("unique non-NA values = %s",
                paste(unique(v[!is.na(v)]), collapse = ","))
      })

# Non-square cells must not be silently squared off.
r_nonsq <- rast(base_ext, nrows = 10, ncols = 40); values(r_nonsq) <- seq_len(400)
check("non-square cells, res preserved",
      apply_pai_raster(model_ok, r_nonsq), expect = "ok", section = sec,
      probe = function(r) sprintf("res in (%.3f, %.3f) -> out (%.3f, %.3f)",
                                  res(r_nonsq)[1], res(r_nonsq)[2],
                                  res(r)[1], res(r)[2]))

report(sec)
```

## 3.2 Categorical data and interpolation method

The case that motivated this audit: a binary mask resampled with `bilinear`
gains values that exist nowhere in the source.

```{r sec-rast-cat}
sec <- "3.2 Categorical rasters"

r_bin <- mk_rast(60, 60, vals = rep(c(0, 1), length.out = 3600))
r_cls <- mk_rast(60, 60, vals = rep(1:5, length.out = 3600))

nunique <- function(r) {
  v <- values(r)[, 1]; length(unique(v[!is.na(v)]))
}

check("binary mask + method = 'near' (values preserved)",
      apply_pai_raster(model_ok, r_bin, method = "near"),
      expect = "ok", section = sec,
      probe = function(r) sprintf("distinct values = %d (source 2)", nunique(r)))
check("binary mask + method = 'bilinear' (blends classes)",
      apply_pai_raster(model_ok, r_bin, method = "bilinear"),
      expect = "ok", section = sec,
      probe = function(r) sprintf("distinct values = %d (source 2)", nunique(r)))
check("5-class raster + 'near'",
      apply_pai_raster(model_ok, r_cls, method = "near"),
      expect = "ok", section = sec,
      probe = function(r) sprintf("distinct values = %d (source 5)", nunique(r)))
check("integer classes survive 'near' as integers",
      apply_pai_raster(model_ok, r_cls, method = "near"),
      expect = "ok", section = sec,
      probe = function(r) {
        v <- values(r)[, 1]; v <- v[!is.na(v)]
        sprintf("all integral = %s", all(v == floor(v)))
      })

report(sec)
```

Bilinear resampling of a categorical raster is *not* an error — the function does
exactly what it was asked to. It is a silent modelling mistake, which is why
`apply_pai_raster()` documents `near` as required for discrete data. Worth noting
that nothing in the package detects it for you.

## 3.3 `mesh_step`, extent and resolution

```{r sec-rast-mesh}
sec <- "3.3 mesh_step and extent"

r_mesh <- mk_rast(40, 40)
exact <- apply_pai_raster(model_gam, r_mesh, ext = "auto", mesh_step = NULL)
na_exact <- sum(is.na(values(exact)[, 1]))

for (ms in c(2, 3, 5, 7, 9, 10, 20)) {
  local({
    step <- ms
    check(sprintf("mesh_step = %d vs exact evaluation", step),
          apply_pai_raster(model_gam, r_mesh, ext = "auto", mesh_step = step),
          expect = "ok", section = sec,
          probe = function(r) {
            n <- sum(is.na(values(r)[, 1]))
            sprintf("NA %d vs exact %d (delta %+d), dim %s",
                    n, na_exact, n - na_exact, paste(dim(r), collapse = "x"))
          })
  })
}

check("mesh_step larger than the raster itself",
      apply_pai_raster(model_gam, mk_rast(8, 8), mesh_step = 50),
      expect = "ok", section = sec,
      probe = function(r) sprintf("dim %s, non-NA = %d",
                                  paste(dim(r), collapse = "x"),
                                  sum(!is.na(values(r)[, 1]))))

check("explicit numeric ext, res = NULL (must preserve resolution)",
      apply_pai_raster(model_ok, r_nonsq,
                       ext = c(hull["xmin"] - 50, hull["xmax"] + 50,
                               hull["ymin"] - 50, hull["ymax"] + 50)),
      expect = "ok", section = sec,
      probe = function(r) sprintf("res in (%.3f, %.3f) -> out (%.3f, %.3f)",
                                  res(r_nonsq)[1], res(r_nonsq)[2],
                                  res(r)[1], res(r)[2]))

check("explicit ext with explicit res",
      apply_pai_raster(model_ok, r_mesh, ext = c(0, 1000, 0, 1000), res = 25),
      expect = "ok", section = sec,
      probe = function(r) sprintf("res = %.2f, dim %s", res(r)[1],
                                  paste(dim(r), collapse = "x")))

check("ext = 'auto' on an identity model (extent should barely move)",
      apply_pai_raster(model_ident, r_mesh, ext = "auto"),
      expect = "ok", section = sec,
      probe = function(r) sprintf("xmin %.1f -> %.1f",
                                  as.vector(ext(r_mesh))[1],
                                  as.vector(ext(r))[1]))

report(sec)
```

## 3.4 Inversion solver

```{r sec-rast-solver}
sec <- "3.4 Coordinate inversion"

r_solver <- mk_rast(20, 20)

check("max_iter = 1 with a tight tolerance (should not converge)",
      apply_pai_raster(model_gam, r_solver, max_iter = 1, tol = 1e-12),
      expect = "warning", section = sec)
check("generous budget (should converge quietly)",
      apply_pai_raster(model_gam, r_solver, max_iter = 50, tol = 1e-4),
      expect = "ok", section = sec)
check("fold-inducing model (non-contractive displacement field)",
      apply_pai_raster(model_fold, r_solver, ext = "auto"),
      expect = "warning", section = sec)
check("lambda = 1 (undamped)",
      apply_pai_raster(model_gam, r_solver, lambda = 1),
      expect = "ok", section = sec)
check("lambda out of range", apply_pai_raster(model_gam, r_solver, lambda = 5),
      expect = "error", section = sec)
check("helmert model uses the closed-form inverse, not iteration",
      apply_pai_raster(train_pai_model(gcp_ok, "helmert"), r_solver),
      expect = "ok", section = sec)

report(sec)
```

## 3.5 Multi-band, AOI and out-of-domain rasters

```{r sec-rast-misc}
sec <- "3.5 Multi-band, AOI, domain"

r_multi <- c(mk_rast(20, 20), mk_rast(20, 20, vals = seq_len(400) * 2))
names(r_multi) <- c("band1", "band2")
check("2-band raster (names and band count preserved)",
      apply_pai_raster(model_ok, r_multi), expect = "ok", section = sec,
      probe = function(r) sprintf("nlyr = %d, names = %s",
                                  nlyr(r), paste(names(r), collapse = ",")))

aoi_half <- st_sf(id = 1, geometry = st_sfc(st_polygon(list(rbind(
  c(hull["xmin"], hull["ymin"]),
  c(mean(hull[1:2]), hull["ymin"]),
  c(mean(hull[1:2]), hull["ymax"]),
  c(hull["xmin"], hull["ymax"]),
  c(hull["xmin"], hull["ymin"])
))), crs = crs_l))
check("AOI covering half the raster (raster has a CRS)",
      apply_pai_raster(model_ok, mk_rast(20, 20), aoi = aoi_half),
      expect = "ok", section = sec,
      probe = function(r) sprintf("NA = %d of %d",
                                  sum(is.na(values(r)[, 1])), ncell(r)))

# The AOI carries a CRS but the raster does not, so the reprojection has no
# target. terra's own message surfaces rather than a mapAI-level explanation.
check("AOI with a CRS, raster without one",
      apply_pai_raster(model_ok, mk_rast(20, 20, crs_in = NULL), aoi = aoi_half),
      expect = "error", section = sec)

# Raster entirely outside the control network.
w <- hull["xmax"] - hull["xmin"]
r_outside <- rast(ext(hull["xmax"] + 5 * w, hull["xmax"] + 6 * w,
                      hull["ymin"], hull["ymin"] + w), nrows = 20, ncols = 20)
values(r_outside) <- seq_len(400)
check("raster entirely outside the GCP footprint (pure extrapolation)",
      apply_pai_raster(model_gam, r_outside, ext = "auto"),
      expect = "ok", section = sec,
      probe = function(r) sprintf("non-NA = %d of %d",
                                  sum(!is.na(values(r)[, 1])), ncell(r)))

check("valueless raster template (no data to sample)",
      apply_pai_raster(model_ok, rast(base_ext, nrows = 10, ncols = 10)),
      expect = "error", section = sec)

report(sec)
```

---

# 4. Distortion analysis

```{r sec-dist}
sec <- "4. Distortion analysis"

check("single evaluation point (degenerate span)",
      analyze_distortion(model_ok, data.frame(source_x = 500, source_y = 500)),
      expect = "ok", section = sec,
      probe = function(d) sprintf("a = %.4f, det_J = %.4f, any NaN = %s",
                                  d$a[1], d$det_J[1],
                                  any(is.nan(c(d$a, d$b, d$det_J, d$theta_a)))))

check("collinear evaluation points",
      analyze_distortion(model_ok,
                         data.frame(source_x = rep(500, 5),
                                    source_y = seq(0, 1000, length.out = 5))),
      expect = "ok", section = sec,
      probe = function(d) sprintf("all finite = %s", all(is.finite(d$a))))

check("identity transformation (a = b = 1, no distortion)",
      analyze_distortion(model_ident, gcp_ident[1:5, ]),
      expect = "ok", section = sec,
      probe = function(d) sprintf("a = %.6f, b = %.6f, 2Omega = %.2e deg",
                                  d$a[1], d$b[1], d$max_angular_distortion[1]))

check("fold-inducing model (negative Jacobian expected)",
      analyze_distortion(suppressWarnings(train_pai_model(gcp_fold, "lm")),
                         gcp_fold[1:5, ]),
      expect = "warning", section = sec)

# Raster input with an NA mask: metrics must stay aligned to their cells.
r_holes <- mk_rast(12, 12)
values(r_holes)[c(1:20, 100:120)] <- NA
check("SpatRaster input with NA cells (alignment + mask)",
      analyze_distortion(model_ok, r_holes), expect = "ok", section = sec,
      probe = function(d) sprintf("cells %d (input %d), NA in 'a' = %d (input %d)",
                                  ncell(d), ncell(r_holes),
                                  sum(is.na(values(d)[, "a"])),
                                  sum(is.na(values(r_holes)[, 1]))))

check("all-NA SpatRaster input",
      analyze_distortion(model_ok, r_allna), expect = "ok", section = sec,
      probe = function(d) sprintf("layers = %d, non-NA = %d",
                                  nlyr(d), sum(!is.na(values(d)[, "a"]))))

check("reference_scale = 0 rejected",
      analyze_distortion(model_ok, gcp_ok[1:3, ], reference_scale = 0),
      expect = "error", section = sec)

check("reference_scale changes both normalised metrics",
      {
        d1 <- analyze_distortion(model_ok, gcp_ok[1:3, ], reference_scale = 1)
        d2 <- analyze_distortion(model_ok, gcp_ok[1:3, ], reference_scale = 2)
        c(log2 = d1$log2_area_scale[1] - d2$log2_area_scale[1],
          airy = d1$airy_kavrayskiy[1] - d2$airy_kavrayskiy[1])
      },
      expect = "ok", section = sec,
      probe = function(v) sprintf("log2 delta = %.3f (expect 2), airy changed = %s",
                                  v["log2"], abs(v["airy"]) > 1e-9))

check("check_tin_inversion() on a non-TIN model",
      check_tin_inversion(model_gam), expect = "warning", section = sec)
check("check_tin_inversion() on a TIN model",
      check_tin_inversion(model_tin), expect = "ok", section = sec)

report(sec)
```

---

# 5. Cross-validation

```{r sec-cv}
sec <- "5. Cross-validation"

gcp_cv <- make_gcp(12, warp = "smooth", noise = 2)

check("k_folds greater than the number of GCPs",
      assess_pai_model(gcp_cv, "lm", validation_type = "random",
                       k_folds = nrow(gcp_cv) + 5),
      expect = "error", section = sec)

check("spatial_block with default block sizing",
      assess_pai_model(gcp_cv, "lm", validation_type = "spatial_block",
                       k_folds = 5, seed = 1),
      expect = "ok", section = sec,
      probe = function(a) sprintf("fold sizes %s, RMSE %.2f",
                                  paste(a$details$n_test_per_fold, collapse = "/"),
                                  a$summary$Mean_RMSE_2D))

check("block_size coarser than the GCP extent (1 block)",
      assess_pai_model(gcp_cv, "lm", validation_type = "spatial_block",
                       k_folds = 5, block_size = 1e6, seed = 1),
      expect = "error", section = sec)

check("block_size giving fewer blocks than k_folds",
      assess_pai_model(gcp_cv, "lm", validation_type = "spatial_block",
                       k_folds = 20, block_size = 400, seed = 1),
      expect = "warning", section = sec)

check("buffer wide enough to empty the training set",
      assess_pai_model(gcp_cv, "lm", validation_type = "spatial_buffered",
                       k_folds = 5, buffer_dist = 1e5, seed = 1),
      expect = "error", section = sec)

check("modest buffer",
      assess_pai_model(gcp_cv, "lm", validation_type = "spatial_buffered",
                       k_folds = 5, buffer_dist = 80, seed = 1),
      expect = "ok", section = sec,
      probe = function(a) sprintf("train sizes %s",
                                  paste(a$details$n_train_per_fold, collapse = "/")))

# An L-shaped footprint leaves many grid cells empty.
gcp_L <- gcp_cv[gcp_cv$source_x < 400 | gcp_cv$source_y < 400, ]
class(gcp_L) <- c("gcp", "data.frame")
check(sprintf("L-shaped GCP footprint (n = %d), spatial_block", nrow(gcp_L)),
      assess_pai_model(gcp_L, "lm", validation_type = "spatial_block",
                       k_folds = 5, seed = 1),
      expect = "ok", section = sec,
      probe = function(a) sprintf("fold sizes %s",
                                  paste(a$details$n_test_per_fold, collapse = "/")))

check("stratified with more strata than distinct displacements",
      assess_pai_model(gcp_ident, "lm", validation_type = "stratified",
                       k_folds = 3, n_strata = 10, seed = 1),
      expect = "error", section = sec)

check("reported metric equals the pooled residual RMSE",
      assess_pai_model(gcp_cv, "lm", validation_type = "spatial", k_folds = 4,
                       seed = 1),
      expect = "ok", section = sec,
      probe = function(a) {
        p <- a$predictions
        pooled <- sqrt(mean((p$true_dx - p$pred_dx)^2 +
                            (p$true_dy - p$pred_dy)^2))
        sprintf("reported %.4f, pooled %.4f, match = %s",
                a$summary$Mean_RMSE_2D, pooled,
                isTRUE(all.equal(a$summary$Mean_RMSE_2D, pooled)))
      })

# A model whose predict() emits NA must not silently delete a fold.
na_model <- list(
  label = "NA Emitter", modelType = "bivariate", library = NULL,
  fit = function(dat, ...) list(),
  predict = function(modelFit, newdata, ...) {
    n <- nrow(newdata)
    dx <- rep(0, n); dy <- rep(0, n); dx[1] <- NA_real_
    data.frame(dx = dx, dy = dy)
  }
)
check("model emitting NA predictions",
      assess_pai_model(gcp_cv, na_model, validation_type = "random",
                       k_folds = 4, seed = 1),
      expect = "warning", section = sec)

check("assess_pai_model leaves the caller's RNG untouched",
      {
        set.seed(4242); before <- .Random.seed
        invisible(assess_pai_model(gcp_cv, "lm", validation_type = "random",
                                   k_folds = 3, seed = 9))
        identical(.Random.seed, before)
      },
      expect = "ok", section = sec,
      probe = function(v) sprintf("RNG preserved = %s", v))

report(sec)
```

---

# Summary

```{r summary}
all_rows <- do.call(rbind, cc$rows)

knitr::kable(as.data.frame(table(
  Section = all_rows$section, Verdict = all_rows$verdict
)) |> subset(Freq > 0), row.names = FALSE)
```

Cases whose behaviour differs from what was expected:

```{r summary-flagged}
flagged <- all_rows[all_rows$verdict != "as expected",
                    c("section", "case", "expected", "observed", "verdict", "detail")]
rownames(flagged) <- NULL
if (nrow(flagged) == 0) {
  cat("None: every case behaved as expected.\n")
} else {
  knitr::kable(flagged)
}
```

`SILENT` is the verdict to care about: a case that was expected to warn or error
but returned quietly. `extra warning` is usually benign — the package was more
talkative than required.

## What this document found, and what was done about it

The first run of this report surfaced two defects and one judgement call. The
first two have since been fixed, and the cases that exposed them are now
regression tests in `test-pai_model_methods.R`. They are recorded here because
the report is only worth keeping if it says what it caught.

### 1. `gam_biv` could not fit a zero-variance displacement field — fixed

A map offset by a constant shift is an ordinary input: `dx = 25`, `dy = -10`
everywhere. `helmert` recovered it exactly and `lm` and `tps` fitted it without
complaint, but `gam_biv` died with

```
Underlying error: NA/NaN/Inf in foreign function call (arg 1)
```

`mgcv::mvn(d = 2)` estimates a full 2x2 residual covariance, which is singular
when a component has no variance. The same failure appeared for an exact identity
warp, for a uniform `1e-9` displacement, and whenever *either* component was
constant regardless of the other — so it was any constant field, not a knife-edge
case. The message named a C argument rather than the modelling situation, and the
scenario it described is precisely "this map needs no correction, or only a rigid
shift".

`gam_biv` now checks the relative standard deviation of each component up front
and refuses with a message naming `helmert` (which recovers a pure translation
exactly) and `lm`. Because mgcv's own numerical tolerance sits at a threshold
that cannot be predicted from the data, the `mgcv::gam()` call is additionally
wrapped so that a singular-covariance failure is re-raised with the same
explanation rather than the raw C message.

### 2. `train_pai_model()` did not validate the control network — fixed

`assess_pai_model()` rejected `NA` control points; `train_pai_model()` ran no
equivalent check, so the two entry points disagreed about usable data:

| Input | was | now |
|---|---|---|
| `NA` in `dx` | fitted on n-1 GCPs, silently | errors, naming the count |
| 1 GCP | model with 4 of 6 coefficients `NA` | errors |
| collinear GCPs | model with 2 of 6 coefficients `NA` | errors, suggests `helmert` |
| co-located GCPs | model with 4 of 6 coefficients `NA` | errors |

A `pai_model` with `NA` coefficients predicts `NA`, so the failure used to surface
later at `transform_map()` or `apply_pai_raster()` time, detached from its cause.
Training was also inconsistent with itself: `helmert` refused co-located points
while `lm` accepted them.

Both entry points now call a shared `validate_gcp_network()`. The checks are
method-aware, because the requirements genuinely differ: `helmert` is a rigid
4-parameter transform that needs only 2 points and stays estimable from collinear
ones, while anything with a bivariate linear term needs a rank-2 point set.
Custom plugins get the universal checks (`NA`, non-finite, co-location) but no
count or rank assumption, since their parameter count is unknown.

### 3. Missing CRS passes without comment — open

`assert_projected_crs()` warns for geographic coordinates but accepts `NA`. A map
with no CRS at all is arguably the more dangerous input, since nothing downstream
can detect a unit mismatch. This one is a judgement call rather than a defect.

### What held up

Everything else. Worth recording explicitly, because these are the cases that
have historically gone wrong in this package:

- All seven OGC geometry types round-trip, interior rings survive as interior
  rings, and mixed-geometry layers still get `area_new`.
- Empty, zero-length, zero-area and zero-row inputs are handled or refused
  cleanly rather than producing corrupt geometry.
- `mesh_step` loses no cells at any step size tested (2, 3, 5, 7, 9, 10, 20),
  including the non-dividing ones.
- Non-square cell sizes survive both `ext = "auto"` and an explicit extent.
- Raster distortion metrics stay aligned to their cells and reproduce the input
  `NA` mask exactly.
- Degenerate distortion inputs (one point, collinear points) return finite
  metrics instead of `NaN`.
- The inversion solver warns on non-convergence and on a non-contractive field
  rather than returning quietly.
- Cross-validation refuses a buffer that would empty the training set, reduces
  `k` when too few blocks are occupied, warns on `NA` predictions, and leaves the
  caller's RNG stream untouched.

```{r sessioninfo}
sessionInfo()
```
