---
title: "Custom Model Training with mapAI"
author: "Generated Vignette"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette demonstrates how to create and evaluate a custom model using the `mapAI` package. We will use the `create_demo_data` function to generate sample data, train a bivariate GAM model, and define a custom model for comparison.

## Generating Demo Data

First, we generate demo data using the `create_demo_data` function:

```{r generate-demo-data}
library(mapAI)

demo_data <- create_demo_data()
```

## Training a Bivariate GAM Model

We train a bivariate GAM model using the `train_pai_model` function:

```{r train-gam-model}
# fit a bivariate GAM model
gam_model <- train_pai_model(gcp_data = demo_data$gcp, method = "gam_biv")
```

## Defining a Custom Model

Next, we define a custom model. This model uses the `mgcv` package to fit a generalized additive model (GAM) with restricted maximum likelihood (REML):

```{r define-custom-model}
# create a custom model
custom_model <- list(
  label = "GAM Model using REML",
  library = "mgcv",
  modelType = "univariate",
  fit = function(x, y, ...) {
    dat <- cbind(y, x)
    names(dat)[1] <- "outcome"

    mgcv::gam(outcome ~ s(source_x) + s(source_y),
            method = "REML",
            data = dat,
            ...)
  },
  predict = function(modelFit, newdata, ...) {
    stats::predict(modelFit, newdata = newdata, ...)
  }
)
```

## Training the Custom Model

We use the `train_pai_model` function to train the custom model:

```{r train-custom-model}
# fit the custom model
custom_fitted <- train_pai_model(gcp_data = demo_data$gcp, method = custom_model)
```

## Comparing Results

Finally, we compare the results of the bivariate GAM model and the custom model using cross-validation:

```{r compare-results}
# compare results
cv_gam <- cv_pai_model(demo_data$gcp, "gam_biv", seed = 1, k = 5)
cv_custom <- cv_pai_model(demo_data$gcp, custom_model, seed = 1, k = 5)
print(cv_gam)
print(cv_custom)
```

This vignette demonstrates the flexibility of the `mapAI` package in training and evaluating custom models.
