---
title: "Custom Model Training with mapAI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Model Training with mapAI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette demonstrates how to define, tune, train, and evaluate custom models using the `mapAI` package. We illustrate how users can incorporate virtually any regression algorithm into `mapAI` by implementing three popular machine learning models:

1. **Support Vector Regression (SVR)** via `e1071`
2. **Neural Network (NN)** via `nnet`
3. **Random Forest (RF)** via `ranger`

### Rigorous Machine Learning Evaluation

To avoid data leakage and optimism bias, we follow standard ML evaluation practices:
- **Data Splitting**: We split the ground control points (GCPs) into an 80% training set (`train_gcp`) and a 20% held-out test set (`test_gcp`).
- **Hyperparameter Tuning on Training Data**: Cross-validation (`cv_pai_model`) is performed exclusively on the training set to identify optimal hyperparameters.
- **Unbiased Test Set Evaluation**: Final models are fitted on the training set and evaluated on the unseen test set, reporting true out-of-sample 2D RMSE.
- **Raster Rectification & Distortion Analysis**: The trained models are applied to correct a continuous raster map and evaluate spatial distortion fields across the domain.

## Generating Demo Data and Train/Test Split

We generate demo data using `create_demo_data(raster = TRUE)` to obtain both homologous control points (`gcp`) and an in-memory `terra` raster representing the distorted map grid:

```{r generate-demo-data}
library(mapAI)
library(terra)

demo_data <- create_demo_data(raster = TRUE, seed = 42)

# Partition GCPs into an 80% training set and a 20% held-out test set
set.seed(42)
n_pts <- nrow(demo_data$gcp)
train_idx <- sample(seq_len(n_pts), size = floor(0.8 * n_pts))

train_gcp <- demo_data$gcp[train_idx, ]
test_gcp  <- demo_data$gcp[-train_idx, ]

cat(sprintf("Training points: %d | Held-out test points: %d\n", nrow(train_gcp), nrow(test_gcp)))
```

## Defining Custom Models

In `mapAI`, a custom model is defined as a list containing the following elements:

- `label`: A descriptive name for the model.
- `library`: Name of the package (or packages) required by the model.
- `modelType`: Either `"univariate"` (fitting separate models for `dx` and `dy`) or `"bivariate"` (fitting both simultaneously).
- `fit`: A function with arguments `x`, `y`, and `...` that trains the model on source coordinates `x` and target displacement `y`.
- `predict`: A function with arguments `modelFit`, `newdata`, and `...` that returns predicted displacement values.

### Custom Model 1: Support Vector Regression (SVR)

```{r define-svr-model}
# Support Vector Regression custom model
svr_model <- list(
  label = "Support Vector Regression",
  library = "e1071",
  modelType = "univariate",
  fit = function(x, y, cost = 1, ...) {
    dat <- cbind(outcome = y, x)
    e1071::svm(outcome ~ source_x + source_y, data = dat, cost = cost, ...)
  },
  predict = function(modelFit, newdata, ...) {
    as.vector(stats::predict(modelFit, newdata = newdata, ...))
  }
)
```

### Custom Model 2: Neural Network

```{r define-nn-model}
# Neural Network custom model
nn_model <- list(
  label = "Neural Network",
  library = "nnet",
  modelType = "univariate",
  fit = function(x, y, size = 5, linout = TRUE, trace = FALSE, ...) {
    dat <- cbind(outcome = y, x)
    nnet::nnet(
      outcome ~ source_x + source_y,
      data = dat,
      size = size,
      linout = linout,
      trace = trace,
      ...
    )
  },
  predict = function(modelFit, newdata, ...) {
    as.vector(stats::predict(modelFit, newdata = newdata, ...))
  }
)
```

### Custom Model 3: Random Forest

```{r define-rf-model}
# Random Forest custom model
rf_model <- list(
  label = "Random Forest",
  library = "ranger",
  modelType = "univariate",
  fit = function(x, y, num.trees = 100, ...) {
    dat <- cbind(outcome = y, x)
    ranger::ranger(formula = outcome ~ ., data = dat, num.trees = num.trees, ...)
  },
  predict = function(modelFit, newdata, ...) {
    stats::predict(modelFit, data = newdata, ...)$predictions
  }
)
```

## Hyperparameter Tuning on Training Data

Because both `train_pai_model()` and `cv_pai_model()` accept `...` and forward extra arguments directly to the custom model's `fit()` function, we can evaluate candidate hyperparameters using 5-fold cross-validation strictly on the training partition.

### Tuning SVR Regularization (`cost`)

```{r tune-svr}
costs <- c(0.1, 1, 10, 50)

svr_cv_results <- vapply(costs, function(c_val) {
  cv <- cv_pai_model(
    gcp_data = train_gcp,
    pai_method = svr_model,
    cost = c_val,
    seed = 1,
    k_folds = 5
  )
  cv$summary$Mean_RMSE_2D
}, numeric(1))

best_cost <- costs[which.min(svr_cv_results)]
cat(sprintf("Optimal SVR cost: %g (CV RMSE: %.3f)\n", best_cost, min(svr_cv_results)))
```

### Tuning Neural Network Hidden Units (`size`)

```{r tune-nn}
hidden_sizes <- c(3, 5, 8)

nn_cv_results <- vapply(hidden_sizes, function(s) {
  cv <- cv_pai_model(
    gcp_data = train_gcp,
    pai_method = nn_model,
    size = s,
    seed = 1,
    k_folds = 5
  )
  cv$summary$Mean_RMSE_2D
}, numeric(1))

best_size <- hidden_sizes[which.min(nn_cv_results)]
cat(sprintf("Optimal NN size: %d (CV RMSE: %.3f)\n", best_size, min(nn_cv_results)))
```

### Tuning Random Forest Ensemble Size (`num.trees`)

```{r tune-rf}
tree_counts <- c(50, 100, 200)

rf_cv_results <- vapply(tree_counts, function(t) {
  cv <- cv_pai_model(
    gcp_data = train_gcp,
    pai_method = rf_model,
    num.trees = t,
    seed = 1,
    k_folds = 5
  )
  cv$summary$Mean_RMSE_2D
}, numeric(1))

best_trees <- tree_counts[which.min(rf_cv_results)]
cat(sprintf("Optimal RF num.trees: %d (CV RMSE: %.3f)\n", best_trees, min(rf_cv_results)))
```

## Training Final Models on Training Data

We train the final models on the entire training set using their respective optimal hyperparameters:

```{r train-final-models}
svr_fitted <- train_pai_model(train_gcp, method = svr_model, cost = best_cost)
nn_fitted  <- train_pai_model(train_gcp, method = nn_model, size = best_size)
rf_fitted  <- train_pai_model(train_gcp, method = rf_model, num.trees = best_trees)
```

## Out-of-Sample Evaluation on the Held-Out Test Set

Now we assess the true generalization accuracy of each model by predicting on the unseen `test_gcp` data points. This guarantees an unbiased performance comparison:

```{r evaluate-test-set}
# Generate predictions on unseen test set
pred_svr <- predict(svr_fitted, newdata = test_gcp)
pred_nn  <- predict(nn_fitted, newdata = test_gcp)
pred_rf  <- predict(rf_fitted, newdata = test_gcp)

# Compute 2D RMSE
calc_rmse_2d <- function(actual, predicted) {
  sqrt(mean((actual$dx - predicted$dx)^2 + (actual$dy - predicted$dy)^2))
}

# Baseline error (untransformed test points)
base_rmse <- sqrt(mean(test_gcp$dx^2 + test_gcp$dy^2))

# Model test set errors
test_rmse_svr <- calc_rmse_2d(test_gcp, pred_svr)
test_rmse_nn  <- calc_rmse_2d(test_gcp, pred_nn)
test_rmse_rf  <- calc_rmse_2d(test_gcp, pred_rf)

# Compile comparison table
comparison_table <- data.frame(
  Model = c("Baseline (Uncorrected)", "Support Vector Regression", "Neural Network", "Random Forest"),
  Optimal_Param = c("-", sprintf("cost = %g", best_cost), sprintf("size = %d", best_size), sprintf("num.trees = %d", best_trees)),
  CV_RMSE_Train = c(NA, round(min(svr_cv_results), 3), round(min(nn_cv_results), 3), round(min(rf_cv_results), 3)),
  Test_RMSE = round(c(base_rmse, test_rmse_svr, test_rmse_nn, test_rmse_rf), 3),
  Improvement_Pct = round(c(
    0,
    (base_rmse - test_rmse_svr) / base_rmse * 100,
    (base_rmse - test_rmse_nn) / base_rmse * 100,
    (base_rmse - test_rmse_rf) / base_rmse * 100
  ), 1)
)

knitr::kable(
  comparison_table,
  caption = "Table: Out-of-sample test set performance compared to training CV results."
)
```

## Comparing Corrected Raster Maps

We can apply each trained model to the distorted raster map (`demo_data$raster`) using `transform_map()`. This performs inverse coordinate mapping and resamples the raster pixel values into the corrected spatial frame:

```{r transform-rasters, fig.width=10, fig.height=6}
# Apply trained models to correct the raster map
svr_raster <- transform_map(svr_fitted, demo_data$raster)
nn_raster  <- transform_map(nn_fitted, demo_data$raster)
rf_raster  <- transform_map(rf_fitted, demo_data$raster)

# Compare original distorted raster and all three model corrections
plot(
  c(demo_data$raster, svr_raster, nn_raster, rf_raster),
  main = c("Original Distorted Map", "SVR Corrected Map", "NN Corrected Map", "RF Corrected Map"),
  col = c("white", "black"),
  legend = FALSE
)
```

## Comparing Model Distortions on Rasters

To evaluate how each model alters geometry across continuous space, we pass the raster to `analyze_distortion()`. This calculates differential distortion metrics based on Tissot's indicatrix theory across all raster cells, returning a multi-layer `terra::SpatRaster`:

```{r analyze-distortions}
# Calculate distortion metrics across the raster
dist_svr <- analyze_distortion(svr_fitted, newdata = demo_data$raster)
dist_nn  <- analyze_distortion(nn_fitted, newdata = demo_data$raster)
dist_rf  <- analyze_distortion(rf_fitted, newdata = demo_data$raster)

# Summarize area scale and maximum shear distortion
cat("=== SVR Distortion Summary ===\n")
summary(dist_svr[[c("area_scale", "max_shear")]])

cat("=== Neural Network Distortion Summary ===\n")
summary(dist_nn[[c("area_scale", "max_shear")]])

cat("=== Random Forest Distortion Summary ===\n")
summary(dist_rf[[c("area_scale", "max_shear")]])
```

### Visualizing Distortion Surfaces

We can plot and compare the distortion fields of the three models:

```{r plot-distortions, fig.width=10, fig.height=6}
# Compare area scale distortion surfaces
plot(
  c(dist_svr$area_scale, dist_nn$area_scale, dist_rf$area_scale),
  main = c("SVR: Area Scale", "NN: Area Scale", "RF: Area Scale")
)

# Compare maximum shear distortion surfaces (degrees)
plot(
  c(dist_svr$max_shear, dist_nn$max_shear, dist_rf$max_shear),
  main = c("SVR: Max Shear (deg)", "NN: Max Shear (deg)", "RF: Max Shear (deg)")
)
```

## Conclusion

This vignette demonstrates an end-to-end machine learning workflow for custom models in `mapAI`:
- **Custom Model Flexibility**: Support Vector Regression, Neural Networks, and Random Forests can be seamlessly integrated into `mapAI`.
- **Methodological Rigor**: Preventing data leakage by tuning hyperparameters exclusively on training data and reporting unbiased generalization performance on held-out test data.
- **Multimodal Evaluation**: Evaluating spatial transformations through cross-validation, held-out test sets, raster map rectification, and continuous Tissot distortion analysis.



