---
title: "Custom Model Training with mapAI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Model Training with mapAI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette demonstrates how to define, tune, train, and evaluate custom models using the `mapAI` package. We illustrate how users can incorporate virtually any regression algorithm into `mapAI` by implementing three popular machine learning models:

1. **Support Vector Regression (SVR)** via `e1071`
2. **Neural Network (NN)** via `nnet`
3. **Random Forest (RF)** via `ranger`

### Rigorous Machine Learning Evaluation

To avoid data leakage and optimism bias, we follow standard ML evaluation practices:
- **Data Splitting**: We split the ground control points (GCPs) into an 80% training set (`train_gcp`) and a 20% held-out test set (`test_gcp`).
- **Hyperparameter Tuning on Training Data**: Cross-validation (`cv_pai_model`) is performed exclusively on the training set to identify optimal hyperparameters.
- **Held-Out Test Set Evaluation**: Final models are fitted on the training set and evaluated on the unseen test set, reporting out-of-sample 2D RMSE. Note the split below is random, not spatial, so the reported figure is unbiased with respect to hyperparameter-selection leakage but still optimistic for spatial generalization (see the capabilities vignette).
- **Raster Rectification & Distortion Analysis**: The trained models are applied to correct a continuous raster map and evaluate spatial distortion fields across the domain.

## Generating Demo Data and Train/Test Split

We generate demo data using `create_demo_data(raster = TRUE)` to obtain both homologous control points (`gcp`) and an in-memory `terra` raster representing the distorted map grid:

```{r generate-demo-data}
library(mapAI)
library(terra)

demo_data <- create_demo_data(raster = TRUE, seed = 42)

# Partition GCPs into an 80% training set and a 20% held-out test set
set.seed(42)
n_pts <- nrow(demo_data$gcp)
train_idx <- sample(seq_len(n_pts), size = floor(0.8 * n_pts))

train_gcp <- demo_data$gcp[train_idx, ]
test_gcp  <- demo_data$gcp[-train_idx, ]

cat(sprintf("Training points: %d | Held-out test points: %d\n", nrow(train_gcp), nrow(test_gcp)))
```

## Defining Custom Models

In `mapAI`, a custom model is defined as a list containing the following elements:

- `label`: A descriptive name for the model.
- `library`: Name of the package (or packages) required by the model.
- `modelType`: Either `"univariate"` (fitting separate models for `dx` and `dy`) or `"bivariate"` (fitting both simultaneously).
- `fit`: the signature depends on `modelType`.
  - For `modelType = "univariate"`, `function(x, y, ...)`, called twice — once for
    `dx` and once for `dy` — where `x` is a data frame of source coordinates and
    `y` the response vector.
  - For `modelType = "bivariate"`, `function(dat, ...)`, called once, where `dat`
    is the **whole GCP data frame** carrying `source_x`, `source_y`, `target_x`,
    `target_y`, `dx` and `dy`.
- `predict`: `function(modelFit, newdata, ...)`. A univariate model returns a
  numeric vector; a bivariate model returns a two-column object whose first
  column is `dx` and second `dy`.

### Custom Model 1: Support Vector Regression (SVR)

```{r define-svr-model}
# Support Vector Regression custom model
svr_model <- list(
  label = "Support Vector Regression",
  library = "e1071",
  modelType = "univariate",
  fit = function(x, y, cost = 1, ...) {
    dat <- cbind(outcome = y, x)
    e1071::svm(outcome ~ source_x + source_y, data = dat, cost = cost, ...)
  },
  predict = function(modelFit, newdata, ...) {
    as.vector(stats::predict(modelFit, newdata = newdata, ...))
  }
)
```

### Custom Model 2: Neural Network

```{r define-nn-model}
# Neural Network custom model
nn_model <- list(
  label = "Neural Network",
  library = "nnet",
  modelType = "univariate",
  fit = function(x, y, size = 5, linout = TRUE, trace = FALSE, ...) {
    dat <- cbind(outcome = y, x)
    nnet::nnet(
      outcome ~ source_x + source_y,
      data = dat,
      size = size,
      linout = linout,
      trace = trace,
      ...
    )
  },
  predict = function(modelFit, newdata, ...) {
    as.vector(stats::predict(modelFit, newdata = newdata, ...))
  }
)
```

### Custom Model 3: Random Forest

```{r define-rf-model}
# Random Forest custom model
rf_model <- list(
  label = "Random Forest",
  library = "ranger",
  modelType = "univariate",
  fit = function(x, y, num.trees = 100, ...) {
    dat <- cbind(outcome = y, x)
    ranger::ranger(formula = outcome ~ ., data = dat, num.trees = num.trees, ...)
  },
  predict = function(modelFit, newdata, ...) {
    stats::predict(modelFit, data = newdata, ...)$predictions
  }
)
```

## Hyperparameter Tuning on Training Data

Because both `train_pai_model()` and `cv_pai_model()` accept `...` and forward extra arguments directly to the custom model's `fit()` function, we can evaluate candidate hyperparameters using 5-fold cross-validation strictly on the training partition.

### Tuning SVR Regularization (`cost`)

```{r tune-svr}
costs <- c(0.1, 1, 10, 50)

svr_cv_results <- vapply(costs, function(c_val) {
  cv <- cv_pai_model(
    gcp_data = train_gcp,
    pai_method = svr_model,
    cost = c_val,
    seed = 1,
    k_folds = 5
  )
  cv$summary$Mean_RMSE_2D
}, numeric(1))

best_cost <- costs[which.min(svr_cv_results)]
cat(sprintf("Optimal SVR cost: %g (CV RMSE: %.3f)\n", best_cost, min(svr_cv_results)))
```

### Tuning Neural Network Hidden Units (`size`)

```{r tune-nn}
hidden_sizes <- c(3, 5, 8)

nn_cv_results <- vapply(hidden_sizes, function(s) {
  cv <- cv_pai_model(
    gcp_data = train_gcp,
    pai_method = nn_model,
    size = s,
    seed = 1,
    k_folds = 5
  )
  cv$summary$Mean_RMSE_2D
}, numeric(1))

best_size <- hidden_sizes[which.min(nn_cv_results)]
cat(sprintf("Optimal NN size: %d (CV RMSE: %.3f)\n", best_size, min(nn_cv_results)))
```

### Tuning Random Forest Ensemble Size (`num.trees`)

```{r tune-rf}
tree_counts <- c(50, 100, 200)

rf_cv_results <- vapply(tree_counts, function(t) {
  cv <- cv_pai_model(
    gcp_data = train_gcp,
    pai_method = rf_model,
    num.trees = t,
    seed = 1,
    k_folds = 5
  )
  cv$summary$Mean_RMSE_2D
}, numeric(1))

best_trees <- tree_counts[which.min(rf_cv_results)]
cat(sprintf("Optimal RF num.trees: %d (CV RMSE: %.3f)\n", best_trees, min(rf_cv_results)))
```

## Training Final Models on Training Data

We train the final models on the entire training set using their respective optimal hyperparameters:

```{r train-final-models}
svr_fitted <- train_pai_model(train_gcp, method = svr_model, cost = best_cost)
nn_fitted  <- train_pai_model(train_gcp, method = nn_model, size = best_size)
rf_fitted  <- train_pai_model(train_gcp, method = rf_model, num.trees = best_trees)
```

## Out-of-Sample Evaluation on the Held-Out Test Set

Now we assess each model on the unseen `test_gcp` points. Because the split is random rather than spatial, nearby training and test points remain correlated, so treat these figures as an upper bound on accuracy for interpolation within the control network, not as spatial generalization performance:

```{r evaluate-test-set}
# Generate predictions on unseen test set
pred_svr <- predict(svr_fitted, newdata = test_gcp)
pred_nn  <- predict(nn_fitted, newdata = test_gcp)
pred_rf  <- predict(rf_fitted, newdata = test_gcp)

# Compute 2D RMSE
calc_rmse_2d <- function(actual, predicted) {
  sqrt(mean((actual$dx - predicted$dx)^2 + (actual$dy - predicted$dy)^2))
}

# Baseline error (untransformed test points)
base_rmse <- sqrt(mean(test_gcp$dx^2 + test_gcp$dy^2))

# Model test set errors
test_rmse_svr <- calc_rmse_2d(test_gcp, pred_svr)
test_rmse_nn  <- calc_rmse_2d(test_gcp, pred_nn)
test_rmse_rf  <- calc_rmse_2d(test_gcp, pred_rf)

# Compile comparison table
comparison_table <- data.frame(
  Model = c("Baseline (Uncorrected)", "Support Vector Regression", "Neural Network", "Random Forest"),
  Optimal_Param = c("-", sprintf("cost = %g", best_cost), sprintf("size = %d", best_size), sprintf("num.trees = %d", best_trees)),
  CV_RMSE_Train = c(NA, round(min(svr_cv_results), 3), round(min(nn_cv_results), 3), round(min(rf_cv_results), 3)),
  Test_RMSE = round(c(base_rmse, test_rmse_svr, test_rmse_nn, test_rmse_rf), 3),
  Improvement_Pct = round(c(
    0,
    (base_rmse - test_rmse_svr) / base_rmse * 100,
    (base_rmse - test_rmse_nn) / base_rmse * 100,
    (base_rmse - test_rmse_rf) / base_rmse * 100
  ), 1)
)

knitr::kable(
  comparison_table,
  caption = "Table: Out-of-sample test set performance compared to training CV results."
)
```

## Comparing Corrected Raster Maps

We can apply each trained model to the distorted raster map (`demo_data$raster`) using `transform_map()`. This performs inverse coordinate mapping and resamples the raster pixel values into the corrected spatial frame:

```{r transform-rasters, fig.width=10, fig.height=6}
# Apply trained models to correct the raster map
svr_raster <- transform_map(svr_fitted, demo_data$raster)
nn_raster  <- transform_map(nn_fitted, demo_data$raster)
rf_raster  <- transform_map(rf_fitted, demo_data$raster)

# Compare original distorted raster and all three model corrections
plot(
  c(demo_data$raster, svr_raster, nn_raster, rf_raster),
  main = c("Original Distorted Map", "SVR Corrected Map", "NN Corrected Map", "RF Corrected Map"),
  col = c("white", "black"),
  legend = FALSE
)
```

## Comparing Model Distortions on Rasters

To evaluate how each model alters geometry across continuous space, we pass the raster to `analyze_distortion()`. This calculates differential distortion metrics based on Tissot's indicatrix theory across all raster cells, returning a multi-layer `terra::SpatRaster`:

```{r analyze-distortions}
# Calculate distortion metrics across the raster
dist_svr <- analyze_distortion(svr_fitted, newdata = demo_data$raster)
dist_nn  <- analyze_distortion(nn_fitted, newdata = demo_data$raster)
dist_rf  <- analyze_distortion(rf_fitted, newdata = demo_data$raster)

# Summarize area scale and maximum angular distortion
cat("=== SVR Distortion Summary ===\n")
summary(dist_svr[[c("area_scale", "max_angular_distortion")]])

cat("=== Neural Network Distortion Summary ===\n")
summary(dist_nn[[c("area_scale", "max_angular_distortion")]])

cat("=== Random Forest Distortion Summary ===\n")
summary(dist_rf[[c("area_scale", "max_angular_distortion")]])
```

### Visualizing Distortion Surfaces

We can plot and compare the distortion fields of the three models:

```{r plot-distortions, fig.width=10, fig.height=6}
# Compare area scale distortion surfaces
plot(
  c(dist_svr$area_scale, dist_nn$area_scale, dist_rf$area_scale),
  main = c("SVR: Area Scale", "NN: Area Scale", "RF: Area Scale")
)

# Compare maximum angular distortion surfaces (degrees)
plot(
  c(dist_svr$max_angular_distortion, dist_nn$max_angular_distortion,
    dist_rf$max_angular_distortion),
  main = c("SVR: Max Angular Distortion (deg)", "NN: Max Angular Distortion (deg)",
           "RF: Max Angular Distortion (deg)")
)
```

## Conclusion

This vignette demonstrates an end-to-end machine learning workflow for custom models in `mapAI`:
- **Custom Model Flexibility**: Support Vector Regression, Neural Networks, and Random Forests can be seamlessly integrated into `mapAI`.
- **Methodological Rigor**: Avoiding hyperparameter-selection leakage by tuning exclusively on training data and reporting performance on held-out test data. For spatial generalization, use one of the spatial cross-validation schemes instead of a random split.
- **Multimodal Evaluation**: Evaluating spatial transformations through cross-validation, held-out test sets, raster map rectification, and continuous Tissot distortion analysis.



