---
title: "Getting Started with mapAI: A Complete Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with mapAI: A Complete Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/start-",
  out.width="100%",
  fig.dim = c(8, 6)
  )
```

This vignette provides a complete, step-by-step workflow for using the `mapAI` package. For users new to the package, this is the best place to start.

We will begin by generating a synthetic dataset that mimics a distorted historical map. This allows us to accurately measure how much our model improves the map's accuracy. The workflow consists of two main parts:

1.  **Positional Correction**: A basic workflow to correct the geometry of a distorted map.
2.  **Advanced Distortion Analysis**: A deeper dive into quantifying and visualizing the nature of the distortion itself using techniques based on Tissot's indicatrix theory.

```{r setup}
# Load the necessary libraries
library(mapAI)
library(ggplot2)
library(sf)
library(dplyr)
library(magrittr) # For the %>% pipe operator
```

## Positional Correction Workflow

### Generate and Load Synthetic Data

First, we use `create_demo_data()` to create our test case. We'll choose the `"complex"` distortion type as it represents a challenging combination of errors. We then load these files into R using the package's reading functions.

```{r create-and-load}
# Create the shapefile and GCPs CSV in a temporary directory
demo_files <- create_demo_data(type = "complex", seed = 123)

# Load the GCPs (homologous points) from the demo file
gcp_data <- read_gcps(gcp_path = demo_files$gcp_path, crs = 3857)

# Load the vector map that needs correction from the demo file
map_to_correct <- read_map(shp_path = demo_files$shp_path, crs = 3857)
```

### Assess Initial Error and Train a Model

We calculate the initial 2D Root Mean Squared Error (RMSE) to get a baseline for the positional error and then we use cross validation to estimate the out-of-sample error of a **Generalized Additive Model (`gam`)**. GAMs are well-suited for capturing the smooth, non-linear distortions in our synthetic data.

```{r train, message=FALSE}
# Calculate the initial error before correction
identity_rmse <- sqrt(mean(gcp_data$dx^2 + gcp_data$dy^2))
print(paste("Initial (Identity) 2D RMSE:", round(identity_rmse, 4)))

# use cross validation to assess a GAM's model accuracy
cv_res <- assess_pai_model(gcp_data,pai_method = "gam")
print(cv_res)
# Train the GAM model using our GCP data
pai_model_gam <- train_pai_model(gcp_data, pai_method = "gam")
```

### Apply Correction and Visualize the Result

We apply our trained model to the map grid and then visualize the result by overlaying the corrected grid on the original.

```{r apply-and-visualize}
# Apply the trained model to the full map
corrected_map <- apply_pai_model(pai_model = pai_model_gam, map = map_to_correct)

# For easy plotting, add a "status" column and combine the maps
map_to_correct$status <- "Original (Distorted)"
corrected_map$status <- "Corrected"
comparison_data <- rbind(map_to_correct[, "status"], corrected_map[, "status"])

# Create the final comparison plot
ggplot(comparison_data) +
  geom_sf(aes(color = status, linetype = status), fill = NA, linewidth = 0.7) +
  scale_color_manual(values = c("Original (Distorted)" = "grey50", "Corrected" = "red")) +
  scale_linetype_manual(values = c("Original (Distorted)" = "dashed", "Corrected" = "solid")) +
  labs(
    title = "Map Correction Comparison",
    subtitle = "Overlay of the original and corrected map grids"
  ) +
  theme_minimal()
```
The plot clearly shows how the warped grid has been adjusted back to a regular shape, demonstrating a successful correction.

---

## Advanced Distortion Analysis

Beyond simple correction, `mapAI` provides powerful tools to quantify and visualize the distortion itself. This is useful for understanding the history of a map's creation or identifying areas of high uncertainty.

### Step 4: Run the Distortion Analysis

We first create a regular grid of **points** for our analysis. Then, we use the `analyze_distortion()` function with our trained `gam_model` to calculate the detailed distortion metrics at each point.

```{r analyze-distortion, message=FALSE}
# Create a regular grid of points for the analysis
analysis_points <- sf::st_make_grid(gcp_data, n = c(25, 25)) %>%
  sf::st_centroid() %>%
  sf::st_sf()

# Run the analysis
distortion_results <- analyze_distortion(pai_model_gam, analysis_points)

# Glimpse the new columns added to our data
glimpse(distortion_results)
```

### Step 5: Visualize Distortion Metrics

Now we can use `plot_distortion_surface()` to visualize these metrics.

#### Areal Distortion (log2σ) and Angular Distortion (2Ω)

These are two of the most fundamental distortion metrics. `log2_area_scale` shows where the map has been stretched or shrunk, while `max_angular_distortion_rad` shows where angles have been deformed (shear).

```{r plot-surfaces}
# Plot for log2(sigma) - areal distortion
plot_area <- plot_distortion_surface(
  distortion_results,
  metric = "log2_area_scale",
  diverging = TRUE
) +
  labs(title = "Areal Distortion (log2σ)")

# Plot for 2*Omega - angular distortion
plot_shear <- plot_distortion_surface(
  distortion_results,
  metric = "max_angular_distortion",
  palette = "magma"
) +
  labs(title = "Angular Distortion (2Ω)")


  plot_shear
  plot_area
```

#### Airy-Kavrayskiy Measure

This is a combined measure of distortion that balances both areal and angular errors. We can calculate it from the `a` and `b` values produced by `analyze_distortion`.

```{r plot-airy}
# Plot the Airy-Kavrayskiy metric
plot_distortion_surface(
  distortion_results,
  metric = "airy_kavrayskiy",
  palette = "cividis"
) +
  labs(title = "Airy-Kavrayskiy Combined Distortion Measure")
```

#### Tissot's Indicatrices

Finally, `plot_indicatrices` provides the richest visualization, showing the distortion ellipses themselves. The shape, size, and orientation of each ellipse represent the local distortion pattern.

```{r plot-indicatrices}
# Plot the indicatrices at the locations of the analysis points
# We use a large scale_factor to make the ellipses clearly visible.
plot_indicatrices(
  distortion_sf = distortion_results,
  scale_factor = 1,
  fill_color = "lightblue",
  border_color = "black"
)
```
This advanced analysis demonstrates how `mapAI` can be used not only to fix historical maps, but to deeply understand their geometric properties.
