---
title: "Model-Agnostic Distortion Diagnostics for Positional Accuracy Improvement"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model-Agnostic Distortion Diagnostics for Positional Accuracy Improvement}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/dist_",
  out.width = "100%",
  fig.dim = c(8, 6),
  message = FALSE,
  warning = FALSE
)
```

## 1. Introduction: The High-Stakes Dilemma of Spatial Data Rectification

Historical cartography, legacy cadastral surveys, scanned map archives, and administrative geospatial datasets provide irreplaceable baselines for analyzing long-term environmental, geomorphic, and land-use change. In historical geographic information systems (HGIS), landscape ecology, and geomorphology, researchers frequently synthesize these historical geometries and imagery with modern basemaps to measure phenomena such as:

* Decadal wetland contraction and drainage,
* Centennial forest fragmentation and agricultural expansion,
* River channel migration and coastal erosion,
* Cadastral boundary evolution and land tenure stability, and
* Historical land-cover transitions extracted from scanned maps and orthomosaics.

To align historical data with modern coordinate reference systems (CRS), **Positional Accuracy Improvement (PAI)** methods are essential. Because historical datasets suffer from non-linear paper shrinkage, composite surveying campaigns, and unrecorded projection variations, global rigid transformations (e.g., Helmert or affine) are rarely sufficient. Consequently, researchers increasingly deploy flexible spatial transformations and machine learning learners, such as Thin Plate Splines (TPS), Generalized Additive Models (GAMs), and Support Vector Machines (SVMs).

### The Danger: Transformation Artifacts Mimic Real Landscape Signals

Flexible non-linear models locally warp coordinate space to eliminate positional discrepancies at known control points. However, in the unmonitored space between control points, flexible transformations risk introducing severe **non-conformal stretching, local areal inflation or deflation, and angular shear**.

Crucially, in a landscape change study, **a spatially-varying transformation artifact is visually and statistically indistinguishable from a true environmental signal**. For instance:

* If an unconstrained model artificially compresses local area by 30% to fit adjacent control points, a researcher measuring parcel or wetland dynamics—whether from vector boundaries or classified raster cells—will erroneously report a **30% real-world habitat loss**.
* If a model introduces local angular shear, orthogonal historical structures, field boundaries, and straight river channels will falsely appear bent or deformed.

### The Blind Spot of Standard Metrics

Conventional spatial validation relies on point-based summary metrics: Root Mean Square Error (RMSE), Mean Absolute Error (MAE), and spatial $k$-fold cross-validation. While these metrics assess how well the model predicts target coordinates at discrete points, **they provide zero information about whether the continuous geospatial fabric between points was compressed, rotated, or sheared**. A model can achieve an impressively low cross-validation RMSE while severely tearing or squishing the continuous geometry of the landscape.

### The Methodological Barrier: Generalizing Beyond Thin Plate Splines

Differential distortion diagnostics based on **Tissot's indicatrix theory** provide the mathematical solution to this dilemma by quantifying local scale factor, areal deformation, and angular shear. However, since the foundational work of Jenny & Hurni (2011) and Claeys Boùùaert et al. (2016), such diagnostics have remained tethered to **closed-form analytical solutions derived strictly for Thin Plate Splines**.

Analytical partial derivatives cannot be readily obtained for black-box or non-parametric machine learning models (e.g., penalized spline GAMs, random forests, neural regressors). The `mapAI` package breaks this 15-year dependency by introducing a **model-agnostic numerical Jacobian estimation engine**. By decoupling distortion analysis from analytical calculus, `mapAI` enables rigorous distortion auditing for *any* differentiable spatial transformation, providing an end-to-end pipeline that accurately rectifies **both vector geometries and raster maps**.

---

## 2. Mathematical Framework: Numerical Jacobian and Tissot's Indicatrix

Let $f: \mathbb{R}^2 \to \mathbb{R}^2$ represent any continuous, trained spatial transformation mapping source coordinates $(x, y)$ to predicted target coordinates $(x', y')$:

$$\begin{pmatrix} x' \\ y' \end{pmatrix} = f(x, y) = \begin{pmatrix} f_x(x, y) \\ f_y(x, y) \end{pmatrix}$$

The local behavior of $f$ around any point $(x, y)$ is governed by its differential linearization, captured by the **Jacobian matrix** $\mathbf{J}$:

$$\mathbf{J}(x, y) = \begin{pmatrix} \frac{\partial x'}{\partial x} & \frac{\partial x'}{\partial y} \\ \frac{\partial y'}{\partial x} & \frac{\partial y'}{\partial y} \end{pmatrix} = \begin{pmatrix} a_{11} & a_{12} \\ a_{21} & a_{22} \end{pmatrix}$$

### Numerical Jacobian Estimation

Rather than requiring an analytical derivation of $f$, `mapAI::analyze_distortion()` computes $\mathbf{J}$ using **central finite differences**:

$$\frac{\partial x'}{\partial x} \approx \frac{f_x(x + h, y) - f_x(x - h, y)}{2h}, \quad \frac{\partial x'}{\partial y} \approx \frac{f_x(x, y + h) - f_x(x, y - h)}{2h}$$

$$\frac{\partial y'}{\partial x} \approx \frac{f_y(x + h, y) - f_y(x - h, y)}{2h}, \quad \frac{\partial y'}{\partial y} \approx \frac{f_y(x, y + h) - f_y(x, y - h)}{2h}$$

where $h$ is an infinitesimal perturbation step (defaulting to $10^{-4}$ in projected coordinate units). Because this formulation queries $f$ strictly as a callable coordinate mapping, it is fully model-agnostic.

### Derivation of Principal Deformation Semi-Axes

Following Tissot's theorem and the singular value decomposition of $\mathbf{J}$, an infinitesimal circle on the source plane deforms into an ellipse on the target plane. The principal semi-major axis $a$ (maximum local scale factor) and semi-minor axis $b$ (minimum local scale factor) are given by:

$$a = \frac{1}{2} \sqrt{(a_{11} + a_{22})^2 + (a_{21} - a_{12})^2} + \frac{1}{2} \sqrt{(a_{11} - a_{22})^2 + (a_{21} + a_{12})^2}$$

$$b = \frac{1}{2} \left| \sqrt{(a_{11} + a_{22})^2 + (a_{21} - a_{12})^2} - \sqrt{(a_{11} - a_{22})^2 + (a_{21} + a_{12})^2} \right|$$

### Operational Distortion Metrics

From $a$ and $b$, `mapAI` computes three core diagnostic metrics across the study domain:

1. **Areal Scale Factor ($\sigma$)**:
   $$\sigma = a \cdot b = \det(\mathbf{J}) = a_{11} a_{22} - a_{12} a_{21}$$
   * Interpretation: $\sigma = 1$ indicates local area preservation (equivalent mapping). Values of $\sigma > 1$ indicate artificial local expansion, while $\sigma < 1$ indicates artificial local compression.
2. **Maximum Angular Distortion ($2\Omega$)**:
   $$2\Omega = 2 \arcsin\left(\frac{a - b}{a + b}\right)$$
   * Interpretation: Quantifies local shear in radians (or degrees). A value of $2\Omega = 0$ indicates local conformality (shape preservation; right angles remain orthogonal).
3. **Airy-Kavrayskiy Combined Metric ($E_{AK}$)**:
   $$E_{AK} = \frac{(a - 1)^2 + (b - 1)^2}{2}$$
   * Interpretation: A composite index penalizing both areal and angular deviation from an ideal isometric transformation ($a = b = 1$).

---

## 3. Comparative Context: Desktop GIS and MapAnalyst

Before running the workflow, it is instructive to compare how spatial adjustment is handled across existing GIS platforms:

| Feature / Capability | ArcGIS Pro (Spatial Adjustment) | QGIS (Vector Bender / Georeferencer) | MapAnalyst | `mapAI` Package |
| :--- | :--- | :--- | :--- | :--- |
| **Data Modalities** | Separate vector rubbersheet / raster georeference | Vector Bender (vectors only) / Georeferencer (rasters only) | Raster scans only (visual diagnostics) | **Unified**: Vectors (`sf`) & Rasters (`terra`) in one pipeline |
| **Adjustment Paradigms** | Piecewise Linear TIN / Natural Neighbor rubbersheet | Piecewise Linear TIN (vectors); TPS / Poly (rasters) | Rigid Affine / Thin Plate Splines | Unified: Geodetic, Regularized Splines (TPS, GAM), and ML |
| **Mathematical Continuity** | $C^0$ (Linear TIN kinks) or $C^1$ | $C^0$ (Linear TIN kinks) | $C^2$ (Analytical TPS) | $C^2$ to $C^\infty$ (Smooth regularized splines) |
| **Point Accuracy Metrics** | Link Table ($dx, dy$) & RMSE | Point residual vectors | Residual vectors & RMSE | Design-based sampling & Spatial cross-validation |
| **Continuous Distortion Diagnostics** | **None** | **None** | **TPS-Only** (Analytical isolines & grids) | **Model-Agnostic** ($\sigma, 2\Omega, E_{AK}$ surfaces via numerical Jacobian) |
| **Operational Output** | Manual interactive GUI editing | Manual interactive GUI editing | **None** (Diagnostic-only; cannot export corrected GIS layers) | Programmatic: `apply_pai_model` (vectors) & `apply_pai_raster` (rasters) |

* **The Desktop GIS Limitation**: ArcGIS Pro and QGIS treat vector adjustment and raster georeferencing as disconnected, interactive GUI operations. Vector rubbersheeting uses Triangulated Irregular Networks (TINs) with $C^0$ derivative discontinuities along triangle edges, introducing artificial kinks into linear features. Furthermore, they report only discrete link residuals, leaving users blind to inter-point fabric distortion.
* **The MapAnalyst Limitation**: MapAnalyst (Jenny & Hurni, 2011) pioneered distortion visualization for historical maps. However, it is **strictly diagnostic**: it cannot transform vector feature layers (`sf` objects) or export rectified raster datasets into analysis-ready GIS layers, and its mathematical formulation is tied exclusively to analytical Thin Plate Splines.

---

## 4. Empirical Workflow: Auditing Cadastral Renewal (1798 Swiss Dataset)

We demonstrate this methodology using the built-in `swiss_cps` dataset: 343 homologous ground control points from Wilhelm Haas's 1798 cadastral map of Basel and Frickthal, aligned with the modern Swiss national coordinate reference system (CH1903 / LV03).

### Setup and Data Inspection

```{r setup}
library(mapAI)
library(sf)
library(terra)
library(dplyr)
library(ggplot2)
library(knitr)

# Load the historical Swiss control points dataset
data(swiss_cps)

# Inspect dataset
head(swiss_cps, 3)
```

We begin by visualizing the raw residual displacement vectors:

```{r displacement_plot}
plot_displacement(swiss_cps) +
  labs(
    title = "Initial Displacement Vectors (Basel and Frickthal, 1798)",
    x = "x (m)",
    y = "y (m)"
  )
```

The displacement vectors reveal systematic, spatially clustered errors, particularly in the southeastern quadrant, confirming that a non-linear approach is required.

---

### The Illusion of Point Accuracy: Model Validation

Before evaluating fabric distortion, we train a **Generalized Additive Model (`gam`)** with penalized thin-plate regression splines and assess its point-level predictive accuracy using both spatial cross-validation and design-based probability sampling:

```{r validation}
# 5-fold spatial cross-validation
cv_spatial <- assess_pai_model(
  swiss_cps,
  pai_method = "gam",
  validation_type = "spatial",
  k_folds = 5,
  seed = 42
)

# Design-based validation using simple random sampling
val_prob <- assess_pai_model(
  swiss_cps,
  pai_method = "gam",
  validation_type = "probability",
  seed = 42
)

# Design-based validation using stratified random sampling
val_strat <- assess_pai_model(
  swiss_cps,
  pai_method = "gam",
  validation_type = "stratified",
  seed = 42
)

# Combine into a summary table
val_summary <- data.frame(
  Strategy = c(
    "Spatial 5-Fold Cross-Validation",
    "Design-Based (Simple Random Sample)",
    "Design-Based (Stratified Sample)"
  ),
  Mean_RMSE_2D_m = round(c(cv_spatial$Mean_RMSE_2D, val_prob$Mean_RMSE_2D, val_strat$Mean_RMSE_2D), 1)
)

kable(val_summary, caption = "Table 1: GAM Point Predictive Accuracy Metrics.")
```

Under standard GIS reporting criteria, the model appears well-behaved: residual RMSE is substantially reduced across all validation schemes. 

---

### The Diagnostic Unmasking: Continuous Distortion Auditing

Now, we train the final model on the full dataset and deploy `analyze_distortion()` across a regular grid over the study area:

```{r model_training}
# Train the final GAM model
final_gam <- train_pai_model(swiss_cps, pai_method = "gam")

# Construct a regular evaluation grid
eval_grid <- sf::st_make_grid(swiss_cps, n = c(40, 40)) %>%
  sf::st_centroid() %>%
  sf::st_sf()

# Perform model-agnostic distortion analysis via numerical Jacobian
distortion_field <- analyze_distortion(final_gam, eval_grid)
```

#### 1. Areal Scale Surface ($\sigma$)

```{r surface_area}
plot_distortion_surface(distortion_field, "area_scale", diverging = TRUE) +
  labs(
    title = expression(paste("Areal Scale Factor (", sigma, ")")),
    subtitle = expression(paste(sigma, " > 1: Expansion; ", sigma, " < 1: Compression")),
    x = "x (m)",
    y = "y (m)",
    fill = expression(sigma)
  )
```

The areal scale surface uncovers severe localized deformation:
* The center-east experiences artificial dilation up to $\sigma = 1.41$ (a **41% artificial expansion**).
* The southeast experiences extreme compression down to $\sigma = 0.45$ (a **55% artificial shrinkage**).

#### 2. Maximum Angular Distortion Surface ($2\Omega$)

```{r surface_shear}
plot_distortion_surface(distortion_field, "max_angular_distortion", palette = "magma") +
  labs(
    title = expression(paste("Maximum Angular Shear (2", Omega, ")")),
    subtitle = "Zero indicates conformality; high values reveal non-orthogonal shearing",
    x = "x (m)",
    y = "y (m)",
    fill = "2Ω (rad)"
  )
```

Angular shearing reaches up to $0.86\text{ rad}$ ($49.3^\circ$), concentrated in the southeastern quadrant. This confirms that orthogonal angles in historical structures would be distorted into acute or obtuse angles by the transformation.

#### 3. Airy-Kavrayskiy Combined Metric ($E_{AK}$)

```{r surface_airy}
plot_distortion_surface(distortion_field, "airy_kavrayskiy") +
  labs(
    title = expression(paste("Airy-Kavrayskiy Metric (", E[AK], ")")),
    subtitle = "Combined index of total geometric distortion",
    x = "x (m)",
    y = "y (m)",
    fill = expression(E[AK])
  )
```

#### 4. Tissot Indicatrices at Control Points

We can also inspect Tissot's indicatrices directly at the control point locations:

```{r indicatrices_plot}
distortion_gcps <- analyze_distortion(final_gam, points_to_analyze = swiss_cps)

plot_indicatrices(distortion_gcps, scale_factor = 700) +
  labs(
    title = "Tissot Indicatrices at Ground Control Points",
    subtitle = "Circular = isometric; Eccentric ellipses = severe local shear and scaling",
    x = "x (m)",
    y = "y (m)"
  )
```

In the center and northwest, the indicatrices are nearly circular ($a \approx b \approx 1$), confirming shape and area fidelity. In the southeast, the indicatrices flatten into highly eccentric ellipses aligned with the direction of maximum shear.

#### Numerical Summary of Distortion

```{r distortion_table}
summary_tbl <- sf::st_drop_geometry(distortion_field) %>%
  reframe(
    Metric = c("Areal Scale Factor (σ)", "Max Angular Shear 2Ω (rad)", "Airy-Kavrayskiy Metric (E_AK)"),
    Minimum = round(c(min(area_scale), min(max_angular_distortion), min(airy_kavrayskiy)), 2),
    Mean = round(c(mean(area_scale), mean(max_angular_distortion), mean(airy_kavrayskiy)), 2),
    Maximum = round(c(max(area_scale), max(max_angular_distortion), max(airy_kavrayskiy)), 2)
  )

kable(summary_tbl, caption = "Table 2: Continuous Distortion Summary Statistics (GAM Model).")
```

---

## 5. Operational Renewal for Vector and Raster Geospatial Data

Having audited the model for deformation artifacts, `mapAI` provides operational functions to apply the learned transformation to both vector geometries and continuous or discrete raster grids.

### 5.1 Mutating Vector Features: `apply_pai_model`

The `apply_pai_model()` function traverses every vertex of simple feature objects (`POINT`, `LINESTRING`, `POLYGON`, `MULTIPOLYGON`), applying the displacement vector $(dx, dy)$:

```{r apply_vector}
# Create a regular line grid to illustrate geometry transformation
sample_grid <- sf::st_make_grid(swiss_cps, n = c(12, 12)) %>%
  sf::st_cast("MULTILINESTRING") %>%
  sf::st_sf()

# Apply the trained GAM model to vectors
corrected_grid <- apply_pai_model(final_gam, sample_grid)

# Plot comparison
sample_grid$Status <- "Original Grid"
corrected_grid$Status <- "GAM Corrected"
grid_comp <- rbind(sample_grid[, "Status"], corrected_grid[, "Status"])

ggplot(grid_comp) +
  geom_sf(aes(color = Status, linetype = Status), linewidth = 0.6) +
  scale_color_manual(values = c("Original Grid" = "grey50", "GAM Corrected" = "#d95f02")) +
  scale_linetype_manual(values = c("Original Grid" = "dashed", "GAM Corrected" = "solid")) +
  theme_minimal() +
  labs(
    title = "Continuous Fabric Transformation (Vector Grid Warping)",
    x = "x (m)",
    y = "y (m)"
  )
```

We evaluate residual point displacements after model fitting:

```{r residual_plot}
plot_residuals(final_gam, swiss_cps) +
  labs(
    title = "Post-Correction Residual Vectors",
    x = "x (m)",
    y = "y (m)"
  )
```

---

### 5.2 Rectifying Raster Grids: `apply_pai_raster`

While vector features are updated by shifting existing vertices forward ($(x, y) \to (x + dx, y + dy)$), raster geospatial data (e.g., scanned historical maps, aerial orthomosaics, classified land cover, digital elevation models) exists on a rigid, regular grid matrix. 

Applying forward mapping directly to raster pixels produces an irregular point cloud with empty gaps ("holes") in dilated areas and cell collisions in compressed areas. To produce a continuous, regular raster grid, `apply_pai_raster()` implements **inverse (backward) mapping**:

1. **Target Grid Definition**: Defines the regular coordinate grid for the target (corrected) raster.
2. **Model Inversion**: Projects each target cell center backwards through the transformation to find its corresponding location in the source (distorted) image. For non-linear models (`gam`, `tps`, `svmRadial`), it employs a rapid fixed-point inversion algorithm that converges to sub-millimeter precision in two iterations.
3. **Resampling**: Resamples pixel values from the source image using either `"bilinear"` interpolation (for continuous surfaces such as scanned map scans, imagery, and DEMs) or `"near"` / `"simple"` nearest-neighbor assignment (essential for categorical rasters like land-use classes or soil types to prevent creating false intermediate class values).
4. **Computational Acceleration (`mesh_step`)**: For multi-gigabyte or high-resolution rasters, setting `mesh_step` (e.g., 10 or 20) computes exact displacements on a subsampled grid and uses high-performance C++ interpolation (`terra::resample`) to reconstruct the fine displacement field, delivering orders-of-magnitude speedups with sub-pixel fidelity.

We demonstrate raster rectification by creating a raster layer across the 1798 Swiss dataset extent:

```{r apply_raster}
# Define raster extent matching swiss_cps bounding box
bb <- sf::st_bbox(swiss_cps)
swiss_ext <- terra::ext(bb["xmin"], bb["xmax"], bb["ymin"], bb["ymax"])

# Create a demo continuous raster (e.g., historical topography / intensity)
historical_raster <- terra::rast(swiss_ext, nrows = 50, ncols = 50, crs = sf::st_crs(swiss_cps)$wkt)
terra::values(historical_raster) <- sin(seq(0, 3 * pi, length.out = terra::ncell(historical_raster))) * 100
names(historical_raster) <- "Elevation_m"

# Rectify the raster using the trained GAM model
corrected_raster <- apply_pai_raster(
  pai_model = final_gam,
  raster = historical_raster,
  method = "bilinear"
)

# Prepare data for comparative visualization
df_orig <- terra::as.data.frame(historical_raster, xy = TRUE)
df_orig$Status <- "Original (Distorted) Raster"
names(df_orig)[3] <- "Value"

df_corr <- terra::as.data.frame(corrected_raster, xy = TRUE)
df_corr$Status <- "GAM Corrected Raster"
names(df_corr)[3] <- "Value"

raster_comp <- rbind(df_orig, df_corr)

ggplot(raster_comp, aes(x = x, y = y, fill = Value)) +
  geom_raster() +
  facet_wrap(~Status) +
  scale_fill_viridis_c(option = "viridis", name = "Value") +
  coord_equal() +
  theme_minimal() +
  labs(
    title = "Raster Map Positional Accuracy Improvement",
    subtitle = "Inverse mapping with bilinear resampling via apply_pai_raster()",
    x = "x (m)",
    y = "y (m)"
  )
```

---

### 5.3 The Geoscientific Takeaway: Avoiding False Change Detection

The southeastern anomaly in the 1798 Basel map is a documented cartographic artifact—likely caused by Wilhelm Haas relying on older, unverified surveys for that quadrant. 

Had a researcher applied an un-audited GAM transformation to measure 200-year land-use change in the southeast:
* **In Vector Analyses**: The model's **55% artificial compression ($\sigma = 0.45$)** would have created severe **phantom wetland contraction or parcel shrinking**.
* **In Raster Analyses**: The backward mapping in the compressed southeast samples pixel values from an artificially large source area, distorting pixel-count-based land-cover statistics.

Standard RMSE metrics never warned the analyst because the GAM fitted the control points accurately. Only continuous, model-agnostic distortion surfaces unmask these localized transformation artifacts.

### 5.4 The 3-Step Distortion-Audited Protocol

To ensure data integrity, we recommend the following protocol when renewing legacy vector or raster geospatial data for scientific reuse:

1. **Candidate Modeling & Point Assessment**: Train candidate models (Helmert, TPS, GAM, ML) and calculate unbiased out-of-sample RMSE using spatial cross-validation or design-based probability sampling.
2. **Fabric Distortion Auditing**: Query the model across a regular evaluation grid using `analyze_distortion()`. Inspect surfaces for areal deformation $\sigma$ and angular shear $2\Omega$. If local scaling exceeds project tolerance thresholds (e.g., $|\sigma - 1| > 0.15$), increase model regularization or collect additional control points.
3. **Safe Geospatial Renewal**: Once continuous fabric integrity is verified, execute `apply_pai_model()` for vector layers or `apply_pai_raster()` for raster maps to generate trustworthy, analysis-ready baselines for downstream environmental change detection.

---

## References

* Claeys Boùùaert, M., De Sloover, L., & De Maeyer, P. (2016). Cartographic deformation analysis of historical maps using thin-plate splines. *Cartography and Geographic Information Science*, 43(4), 356–368.
* Jenny, B., & Hurni, L. (2011). Studying cartographic heritage: Analysis and visualization of geometric distortions. *Computers & Geosciences*, 37(1), 71–81.
* Laskowski, P. C. (1989). The traditional and modern look at Tissot's indicatrix. *The American Cartographer*, 16(2), 123–133.
* Wadoux, A. M. J. C., Brus, D. J., & Heuvelink, G. B. M. (2021). Sampling design for validating digital soil maps: A tutorial. *Geoderma*, 388, 114923.
