---
title: "A Rigorous Methodological and Computational Framework for Positional Accuracy Improvement: The mapAI Package"
author: "Konstantinos Vantas"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{A Rigorous Methodological and Computational Framework for Positional Accuracy Improvement: The mapAI Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  warning = FALSE,
  message = FALSE
)
```

# Executive Summary & Architectural Overview

The **`mapAI`** package provides a statistically rigorous, computationally efficient, and topology-resilient framework for **Positional Accuracy Improvement (PAI)** of spatial datasets in R. Designed for geodesists, GIS professionals, and spatial data scientists, `mapAI` bridges classical geodetic coordinate transformations with modern non-parametric regression, spatial machine learning, and differential geometry diagnostics.

Historically, georeferencing and geometric correction have relied on global polynomial or similarity transformations. While sufficient for simple translations and rigid rotations, such models fail catastrophically when applied to historical cartography, scanned analog cadastre, or aerial imagery subject to complex, spatially continuous, non-linear deformations. 

To address these challenges, `mapAI` establishes a complete end-to-end pipeline:
1. **Algorithmic Diversity**: Implements conformal similarity (Helmert via Ordinary and Total Least Squares), general affine, Thin Plate Splines (TPS), Bivariate Generalized Additive Models (GAM) with Thin Plate Regression Splines, and an extensible interface for arbitrary machine learning algorithms (Support Vector Machines, Random Forests, Neural Networks).
2. **Spatial Data Leakage Prevention**: Features an integrated spatial cross-validation framework supporting six distinct partition topologies (`random`, `spatial`, `spatial_block`, `spatial_buffered`, `probability`, `stratified`) to guard against spatial autocorrelation bias and overfitting.
3. **Topology-Preserving Vector Transformation**: Offers a robust geometry reconstruction engine capable of processing all standard OGC simple feature types (`POINT`, `LINESTRING`, `POLYGON` with interior boundary rings, `MULTIPOINT`, `MULTILINESTRING`, `MULTIPOLYGON`), spatial AOI partitioning, and automated topological repair via `sf::st_make_valid()`.
4. **In-Memory Raster Rectification**: Implements backward mapping with a damped Picard-Mann fixed-point iterative coordinate inversion solver and regular mesh interpolation, executing entirely in RAM via `terra::SpatRaster`.
5. **Differential Distortion & Inversion Diagnostics**: Evaluates local deformation fields using Tissot's Indicatrix, the Cauchy-Green metric tensor, and the signed Jacobian determinant $\det(\mathbf{J})$, automatically flagging topological fold-overs and singular shears.

---

# Theoretical Foundations of Positional Accuracy Improvement

## The Continuum Mechanics Formulation of Map Deformation

Let $\Omega_s \subset \mathbb{R}^2$ represent the coordinate domain of a distorted spatial dataset (the **source** space), and let $\Omega_t \subset \mathbb{R}^2$ represent the true geographic or geodetic reference domain (the **target** space). A set of $n$ homologous Ground Control Points (GCPs) is observed across both domains:

$$\mathcal{D} = \left\{ \left( \mathbf{s}_i, \mathbf{t}_i \right) \right\}_{i=1}^n, \quad \mathbf{s}_i = \begin{pmatrix} x_{s,i} \\ y_{s,i} \end{pmatrix} \in \Omega_s, \quad \mathbf{t}_i = \begin{pmatrix} x_{t,i} \\ y_{t,i} \end{pmatrix} \in \Omega_t$$

The goal of Positional Accuracy Improvement is to estimate a continuous spatial mapping function $\mathbf{f}: \Omega_s \to \Omega_t$ such that:

$$\mathbf{t} = \mathbf{f}(\mathbf{s}) = \mathbf{s} + \mathbf{d}(\mathbf{s})$$

where $\mathbf{d}(\mathbf{s}) = \left( d_x(\mathbf{s}), d_y(\mathbf{s}) \right)^T$ is the continuous 2D displacement field defined by:

$$d_x(\mathbf{s}) = x_t - x_s, \qquad d_y(\mathbf{s}) = y_t - y_s$$

In physical terms, $\mathbf{d}(\mathbf{s})$ encapsulates the superposition of several distinct error processes:
- **Global Datum Misalignment**: Rigid translation, scale divergence, and azimuth orientation discrepancies.
- **Systematic Non-Orthogonal Distortion**: Anisotropic paper grain shrinkage or expansion driven by relative humidity variations.
- **Local Non-Linear Irregularities**: Paper warping, non-uniform scanning lens distortions, plane-table surveying imprecision, and relief displacement in historical surveys (Valente et al., 2021).

---

# Algorithmic Taxonomy & Mathematical Implementations

The `mapAI` package implements a tiered hierarchy of parametric, semi-parametric, and machine learning models.

```{r load-libraries}
library(mapAI)
library(sf)
library(terra)
library(ggplot2)
```

## Conformal Similarity Transformation: Helmert (OLS and TLS)

The 4-parameter Helmert transformation preserves local shape and angles (conformal mapping) by applying a uniform scale factor $\mu > 0$, an orthogonal rotation angle $\theta \in [-\pi, \pi)$, and translations $(t_x, t_y)^T$:

$$\begin{pmatrix} x_t \\ y_t \end{pmatrix} = \begin{pmatrix} c & -d \\ d & c \end{pmatrix} \begin{pmatrix} x_s \\ y_s \end{pmatrix} + \begin{pmatrix} t_x \\ t_y \end{pmatrix}$$

where $c = \mu \cos(\theta)$ and $d = \mu \sin(\theta)$. The physical parameters are uniquely reconstructed via:

$$\mu = \sqrt{c^2 + d^2}, \qquad \theta = \operatorname{atan2}(d, c)$$

### Ordinary Least Squares (OLS)

Under classical OLS, source coordinates $\mathbf{s}_i$ are treated as error-free deterministic constants, with stochastic residuals confined exclusively to target measurements:

$$\min_{c, d, t_x, t_y} \sum_{i=1}^n \left( \left[ c x_{s,i} - d y_{s,i} + t_x - x_{t,i} \right]^2 + \left[ d x_{s,i} + c y_{s,i} + t_y - y_{t,i} \right]^2 \right)$$

Centering coordinates about their centroids $\bar{\mathbf{s}}$ and $\bar{\mathbf{t}}$ yields the closed-form normal equations:

$$c = \frac{\sum_{i=1}^n (u_i X_i + v_i Y_i)}{\sum_{i=1}^n (u_i^2 + v_i^2)}, \qquad d = \frac{\sum_{i=1}^n (u_i Y_i - v_i X_i)}{\sum_{i=1}^n (u_i^2 + v_i^2)}$$

where $u_i = x_{s,i} - \bar{x}_s$, $v_i = y_{s,i} - \bar{y}_s$, $X_i = x_{t,i} - \bar{x}_t$, and $Y_i = y_{t,i} - \bar{y}_t$.

### Total Least Squares (TLS) / Errors-in-Variables

When source coordinates are also subject to measurement errors (e.g., manual digitizing errors on scanned maps), OLS yields biased parameter estimates (attenuation bias). `mapAI` solves this Errors-in-Variables (EIV) problem via **Total Least Squares (TLS)** using Singular Value Decomposition (SVD) (Golub & Van Loan, 2013).

The centered augmented data matrix is formed as:

$$\mathbf{M} = \begin{pmatrix} \mathbf{u} & -\mathbf{v} & \mathbf{X} \\ \mathbf{v} & \mathbf{u} & \mathbf{Y} \end{pmatrix}_{2n \times 3}$$

Computing the thin SVD $\mathbf{M} = \mathbf{U} \mathbf{\Sigma} \mathbf{V}^T$, the right singular vector $\mathbf{v}_3 = (v_{31}, v_{32}, v_{33})^T$ corresponding to the smallest singular value $\sigma_3$ provides the optimal TLS solution:

$$\begin{pmatrix} c \\ d \end{pmatrix} = -\frac{1}{v_{33}} \begin{pmatrix} v_{31} \\ v_{32} \end{pmatrix}$$

The a posteriori reference variance $\hat{\sigma}_0^2$ and standard errors with $\nu = 2n - 4$ degrees of freedom are computed as:

$$\hat{\sigma}_0^2 = \frac{\|\mathbf{r}\|_2^2}{2n - 4}, \qquad \operatorname{SE}(\hat{\mathbf{p}}) = \hat{\sigma}_0 \sqrt{\operatorname{diag}\left( (\mathbf{A}^T \mathbf{A})^{-1} \right)}$$

```{r helmert-example}
# Generate synthetic GCPs with known rigid shift, 5 deg rotation, 1.05 scale
set.seed(42)
n_pts <- 25
xs <- runif(n_pts, 1000, 5000)
ys <- runif(n_pts, 2000, 6000)
theta_true <- 5 * pi / 180
scale_true <- 1.05
tx_true <- 250; ty_true <- -180

xt <- scale_true * (cos(theta_true)*xs - sin(theta_true)*ys) + tx_true + rnorm(n_pts, 0, 0.5)
yt <- scale_true * (sin(theta_true)*xs + cos(theta_true)*ys) + ty_true + rnorm(n_pts, 0, 0.5)

sample_gcp <- read_gcp(source_x = xs, source_y = ys, target_x = xt, target_y = yt)

# Fit OLS and TLS Helmert models
model_ols <- helmert(source_x = sample_gcp$source_x, source_y = sample_gcp$source_y,
                     target_x = sample_gcp$target_x, target_y = sample_gcp$target_y,
                     method = "ols")
model_tls <- helmert(source_x = sample_gcp$source_x, source_y = sample_gcp$source_y,
                     target_x = sample_gcp$target_x, target_y = sample_gcp$target_y,
                     method = "tls")

cat(sprintf("True: Scale=%.4f, Rot=%.2f deg\nOLS:  Scale=%.4f, Rot=%.2f deg (sigma0=%.3f m)\nTLS:  Scale=%.4f, Rot=%.2f deg (sigma0=%.3f m)\n",
            scale_true, theta_true * 180 / pi,
            model_ols$scale, model_ols$theta_deg, model_ols$sigma0,
            model_tls$scale, model_tls$theta_deg, model_tls$sigma0))
```

## General Affine Transformation (`lm`)

The 6-parameter affine model allows for independent scaling along the $x$ and $y$ axes and non-orthogonal shear:

$$\begin{pmatrix} x_t \\ y_t \end{pmatrix} = \begin{pmatrix} a_{11} & a_{12} \\ a_{21} & a_{22} \end{pmatrix} \begin{pmatrix} x_s \\ y_s \end{pmatrix} + \begin{pmatrix} t_x \\ t_y \end{pmatrix}$$

This is fitted via multivariate linear regression (`stats::lm`), capturing anisotropic paper shrinkage between grain and cross-grain directions.

## Thin Plate Splines (`tps`)

The Thin Plate Spline (TPS) minimizes an objective function balancing interpolation fidelity with an idealized physical bending energy of a thin metal sheet (Bookstein, 1989):

$$\mathcal{J}_{\text{TPS}}(f) = \sum_{i=1}^n \|\mathbf{t}_i - f(\mathbf{s}_i)\|^2 + \lambda \iint_{\mathbb{R}^2} \left[ \left(\frac{\partial^2 f}{\partial x^2}\right)^2 + 2\left(\frac{\partial^2 f}{\partial x \partial y}\right)^2 + \left(\frac{\partial^2 f}{\partial y^2}\right)^2 \right] dx dy$$

The smooth mapping decomposes into a global affine component and a linear combination of radial basis functions $\phi(r) = r^2 \ln(r)$:

$$f(\mathbf{s}) = \mathbf{A} \mathbf{s} + \mathbf{b} + \sum_{i=1}^n \mathbf{w}_i \phi(\|\mathbf{s} - \mathbf{s}_i\|)$$

## Bivariate Generalized Additive Models (`gam_biv`)

While classical TPS uses all $n$ data points as knots—yielding an $\mathcal{O}(n^3)$ computational bottleneck—`mapAI` employs **Thin Plate Regression Splines (TPRS)** via `mgcv::gam` (Wood, 2003). The displacement field is modeled as:

$$d_x(\mathbf{s}) = \beta_{0,x} + \sum_{j=1}^k \beta_{j,x} b_j(\mathbf{s}), \qquad d_y(\mathbf{s}) = \beta_{0,y} + \sum_{j=1}^k \beta_{j,y} b_j(\mathbf{s})$$

The smoothing parameters $\boldsymbol{\lambda} = (\lambda_x, \lambda_y)$ are estimated via **Restricted Maximum Likelihood (REML)**. To prevent rank-deficiency on small sample sizes, `mapAI` uses an adaptive basis dimension formula:

$$k = \max\left(3, \; \min\left(29, \; \lfloor 0.6 \times n_{\text{unique}} \rfloor \right)\right)$$

guaranteeing numerical robustness across datasets ranging from sparse GCP sets ($N \approx 10$) to regional surveys ($N > 1000$).

---

# Mitigating Spatial Autocorrelation Bias: Six Validation Schemes

## The Spatial Autocorrelation Fallacy

In classical machine learning, data points are assumed to be Independent and Identically Distributed (IID). Under **Tobler's First Law of Geography**—*"everything is related to everything else, but near things are more related than distant things"* (Tobler, 1970)—spatial residuals exhibit positive autocorrelation:

$$\operatorname{Cov}(\epsilon(\mathbf{s}_i), \epsilon(\mathbf{s}_j)) = C(\|\mathbf{s}_i - \mathbf{s}_j\|) > 0$$

When using standard random cross-validation, training and testing points frequently lie in close geographical proximity. The model can simply memorize local residual patterns rather than learning the generalized deformation field. This creates **spatial data leakage**, resulting in severe optimism bias where reported validation errors are artificially deflated (Roberts et al., 2017).

## The Six Cross-Validation Strategies in `mapAI`

To eliminate spatial data leakage, `mapAI` provides six validation schemes via `cv_pai_model()` and directly inside `train_pai_model(..., cv = TRUE)`:

| Strategy | `cv_args$method` | Splitting Mechanism | Safeguard Against Leakage |
| :--- | :--- | :--- | :--- |
| **Random $k$-Fold** | `"random"` | Uniform random assignment across $k$ folds | None (Baseline IID benchmark) |
| **Spatial Clustered** | `"spatial"` | $k$-Means clustering on spatial coordinates | Disjoint geographic clusters |
| **Spatial Block** | `"spatial_block"` | Regular $n_x \times n_y$ rectangular spatial grid | Independent geographic blocks |
| **Spatial Buffered** | `"spatial_buffered"` | Block holdout with exclusion buffer zone $\delta$ | Complete dead-zone around test fold |
| **Probability Sampling** | `"probability"` | Design-based sampling with coordinate-density weights | Reflects spatial sample design |
| **Stratified Sampling** | `"stratified"` | Stratified by residual displacement magnitude | Balances extreme displacement tails |

## Empirical Benchmark Across All Six Strategies

We demonstrate the empirical behavior of all six validation strategies using the built-in Swiss historical dataset (`basel_frickthal$gcp`), consisting of 343 homologous control points from the 1798 Meyer-Weiss map:

```{r validation-benchmark}
data(basel_frickthal)
swiss_cps <- basel_frickthal$gcp

# Compare all 6 validation strategies on the Bivariate GAM model
methods <- c("random", "spatial", "spatial_block", "spatial_buffered", "probability", "stratified")
results <- list()

set.seed(123)
for (m in methods) {
  cv_cfg <- list(validation_type = m, k_folds = 5)
  if (m == "spatial_buffered") {
    cv_cfg$buffer_dist <- 2000 # 2 km exclusion buffer
  }
  
  mod <- train_pai_model(gcp_data = swiss_cps, method = "gam_biv", cv = cv_cfg)
  results[[m]] <- data.frame(
    Method = m,
    CV_2D_RMSE_m = round(mod$cv_rmse_2d, 3),
    Residual_SD_m = round(mod$cv$summary$SD_RMSE_2D, 3)
  )
}

benchmark_table <- do.call(rbind, results)
knitr::kable(benchmark_table, 
             caption = "Out-of-sample 2D RMSE across the 6 validation schemes for the 1798 Basel-Frickthal map.")
```

### Analysis of the Benchmark Findings
- **Random Cross-Validation (`"random"`)** reports the lowest RMSE because test folds are surrounded by spatially proximate training points.
- **Spatial Block (`"spatial_block"`) and Buffered Block (`"spatial_buffered"`)** report higher, realistic prediction errors because the model is evaluated on unseen geographic regions without relying on local spatial autocorrelation.
- When publishing empirical evaluations or assessing georeferencing quality, practitioners should report spatial block or buffered block metrics to reflect true out-of-sample generalization.

---

# The Vector Transformation Engine & Topological Resilience

## Decomposition and Geometry Reconstruction in `transform_map`

The `transform_map()` function implements a multi-stage vector transformation pipeline:
1. **Coordinate System Verification**: Asserts that inputs possess a projected CRS with linear meter units, issuing clear warnings for unprojected angular coordinates (longitude/latitude).
2. **Feature-by-Feature Vertex Extraction**: Decomposes complex OGC geometries into constituent 2D point arrays:
   - `POINT`: Direct transformation $(x_t, y_t) = \mathbf{f}(x_s, y_s)$.
   - `LINESTRING` & `MULTILINESTRING`: Sequential vertex transformation preserving coordinate order.
   - `POLYGON`: Preserves outer exterior boundaries and nested interior hole boundaries.
   - `MULTIPOLYGON`: Preserves multi-part topological structures.
3. **Batch Model Prediction**: Evaluates $\mathbf{d}(\mathbf{s})$ in vectorized batches, minimizing interpreter overhead.
4. **Area Recalculation**: For areal features (`POLYGON`, `MULTIPOLYGON`), automatically recalculates geometric area using planar metric properties.

## Topological Validity & Automated Repair

Non-linear displacement fields $\mathbf{f}(\mathbf{s})$ can introduce severe topological defects into vector geometries:
- Boundary edge self-intersections (butterfly / bowtie polygons).
- Exterior ring reversals.
- Collapsing of thin polygon slivers.

`transform_map()` integrates automated topological validation and repair:

$$\text{Valid}(\mathcal{G}) = \begin{cases} \text{Keep } \mathcal{G} & \text{if } \texttt{sf::st\_is\_valid}(\mathcal{G}) = \text{TRUE} \\ \texttt{sf::st\_make\_valid}(\mathcal{G}) & \text{if } \texttt{sf::st\_is\_valid}(\mathcal{G}) = \text{FALSE and } \texttt{repair\_topology} = \text{TRUE} \end{cases}$$

```{r vector-transformation-demo}
# Create a synthetic parcel with boundary ring
parcel_poly <- st_polygon(list(matrix(c(
  2610000, 1260000,
  2615000, 1260000,
  2615000, 1265000,
  2610000, 1265000,
  2610000, 1260000
), ncol = 2, byrow = TRUE)))

parcel_sf <- st_sf(parcel_id = "CH-01", geometry = st_sfc(parcel_poly, crs = 2056))

# Fit a TPS model on the Basel GCPs
model_tps <- train_pai_model(gcp_data = swiss_cps, method = "tps")

# Apply vector transformation with automated topology repair
transformed_parcel <- transform_map(
  map = parcel_sf,
  pai_model = model_tps,
  repair_topology = TRUE
)

cat(sprintf("Original Area:    %.2f sq m\nTransformed Area: %.2f sq m\nTopology Valid:   %s\n",
            st_area(parcel_sf),
            transformed_parcel$area_m2,
            st_is_valid(transformed_parcel)))
```

---

# In-Memory Raster Rectification: The Damped Picard Inversion Solver

## Forward Mapping vs. Backward Mapping

Rectifying a continuous raster grid presents a fundamental mathematical dilemma:
- **Forward Mapping**: Moving source pixels $\mathbf{s}_{i,j}$ to target locations $\mathbf{t}_{i,j} = \mathbf{f}(\mathbf{s}_{i,j})$. Because $\mathbf{f}$ is non-linear, target coordinates fall unevenly across continuous space, producing empty "raster holes" (unassigned cells) and overlapping cell collisions.
- **Backward Mapping**: Iterating over regular target grid pixels $\mathbf{t}_{u,v}$ and finding the corresponding source coordinate $\mathbf{s}_{u,v} = \mathbf{f}^{-1}(\mathbf{t}_{u,v})$, followed by bilinear interpolation of pixel values.

However, PAI models are explicitly trained in the **forward direction** ($\mathbf{t} = \mathbf{s} + \mathbf{d}(\mathbf{s})$). For non-linear models like TPS and GAM, no analytical inverse $\mathbf{f}^{-1}$ exists.

## Damped Picard-Mann Iterative Fixed-Point Inversion

To solve the backward mapping problem without requiring explicit model inversion, `apply_pai_raster()` implements a **Damped Picard-Mann Iterative Fixed-Point Coordinate Solver**.

Given a target cell coordinate $\mathbf{t}$, we seek $\mathbf{s}^*$ such that:

$$\mathbf{s}^* + \mathbf{d}(\mathbf{s}^*) = \mathbf{t} \iff \mathbf{s}^* = \mathbf{t} - \mathbf{d}(\mathbf{s}^*)$$

Starting with initial estimate $\mathbf{s}^{(0)} = \mathbf{t} - \mathbf{d}(\mathbf{t})$, the iterative sequence proceeds as:

$$\mathbf{s}^{(k+1)} = \mathbf{s}^{(k)} - \lambda \left[ \mathbf{s}^{(k)} + \mathbf{d}(\mathbf{s}^{(k)}) - \mathbf{t} \right], \quad k = 0, 1, \dots, K_{\max}$$

where $\lambda \in (0, 1]$ is a relaxation damping factor ($\lambda = 0.70$ by default). Iteration terminates when the Euclidean step size falls below tolerance $\epsilon = 10^{-4}$ meters:

$$\|\mathbf{s}^{(k+1)} - \mathbf{s}^{(k)}\|_2 < \epsilon$$

## Subsampled Mesh Bilinear Interpolation

Evaluating the fixed-point solver for every pixel on a 50-megapixel raster is computationally prohibitive. `apply_pai_raster()` constructs a coarse regular mesh across the target domain (e.g., $100 \times 100$ control nodes), solves the iterative inversion exclusively at mesh vertices, and applies fast continuous bilinear interpolation across intermediate pixels.

```{r raster-rectification-demo}
# Create synthetic raster map representing distorted survey map
r_distorted <- rast(nrows = 60, ncols = 60,
                    xmin = 2605000, xmax = 2625000,
                    ymin = 1255000, ymax = 1275000,
                    crs = "EPSG:2056")
values(r_distorted) <- sin(seq(0, 4*pi, length.out = ncell(r_distorted)))

# Rectify raster entirely in memory
r_rectified <- apply_pai_raster(
  raster = r_distorted,
  pai_model = model_tps,
  mesh_step = 2,
  max_iter = 15,
  tol = 1e-4,
  ext = "auto"
)

cat(sprintf("Original Raster Dimensions:  %d x %d\nRectified Raster Dimensions: %d x %d\nIn-Memory Processing:        %s\n",
            nrow(r_distorted), ncol(r_distorted),
            nrow(r_rectified), ncol(r_rectified),
            inMemory(r_rectified)))
```

---

# Differential Geometry & Topological Inversion Diagnostics

## The Metric Tensor and Tissot's Indicatrix

Any continuous map transformation $\mathbf{f}(x, y) = (u(x, y), v(x, y))^T$ can be characterized locally by its **Jacobian matrix** $\mathbf{J}$:

$$\mathbf{J}(x, y) = \begin{pmatrix} \frac{\partial u}{\partial x} & \frac{\partial u}{\partial y} \\ \frac{\partial v}{\partial x} & \frac{\partial v}{\partial y} \end{pmatrix} = \begin{pmatrix} 1 + \frac{\partial d_x}{\partial x} & \frac{\partial d_x}{\partial y} \\ \frac{\partial d_y}{\partial x} & 1 + \frac{\partial d_y}{\partial y} \end{pmatrix}$$

The **Cauchy-Green metric tensor** $\mathbf{G} = \mathbf{J}^T \mathbf{J}$ describes how infinitesimal lengths and angles are distorted. According to Tissot's theorem (Tissot, 1881), an infinitesimal circle in the source domain is mapped to an infinitesimal ellipse (**Tissot's Indicatrix**) in the target domain.

The eigenvalues of $\mathbf{G}$, denoted $\lambda_1 \ge \lambda_2 > 0$, correspond to the squared lengths of the semi-major ($a$) and semi-minor ($b$) axes of Tissot's ellipse:

$$a = \sqrt{\lambda_1}, \qquad b = \sqrt{\lambda_2}$$

From these principal axes, `analyze_distortion()` extracts fundamental deformation invariants:
1. **Area Scale Factor ($s$)**: Ratio of transformed area to original area:
   $$s = a \cdot b = |\det(\mathbf{J})|$$
2. **Maximum Angular Distortion ($2\omega$)**: Maximum angular deviation experienced by orthogonal intersecting lines:
   $$2\omega = 2 \arcsin\left( \frac{a - b}{a + b} \right)$$
3. **Areal Strain ($\sigma$)**: Percentage change in differential surface area:
   $$\sigma = s - 1$$

## Topological Inversion: The Signed Jacobian Determinant

While the classical Tissot formulation considers only the absolute area scale factor $|\det(\mathbf{J})|$, `mapAI` computes the **signed Jacobian determinant**:

$$\det(\mathbf{J}) = \frac{\partial u}{\partial x}\frac{\partial v}{\partial y} - \frac{\partial u}{\partial y}\frac{\partial v}{\partial x}$$

The sign of $\det(\mathbf{J})$ carries vital topological significance:
- **$\det(\mathbf{J}) > 0$**: The transformation is orientation-preserving and locally homeomorphic.
- **$\det(\mathbf{J}) = 0$**: The transformation has collapsed spatial dimensions into a line or point (singular shear).
- **$\det(\mathbf{J}) < 0$**: The transformation has folded the map onto itself (topological fold-over / reflection), breaking geographic reality.

Whenever $\det(\mathbf{J}) \le 0$ occurs within the evaluation grid, `analyze_distortion()` triggers an explicit warning alerting the practitioner to local topological inversion.

```{r distortion-analysis-demo}
# Compute differential distortion across the Swiss GCPs
distortion_res <- analyze_distortion(
  pai_model = model_tps
)

# Print comprehensive summary
summary(distortion_res)
```

Visualizing the spatial distribution of angular distortion, area scale, and displacement vectors reveals where the historical map experienced severe physical strain:

```{r plot-distortion-fields, fig.width=7, fig.height=5}
# Plot spatial displacement vectors and distortion metrics
plot(swiss_cps)
plot(distortion_res, metric = "area_scale")
plot(distortion_res, metric = "max_angular_distortion")
indicatrices(distortion_res)
```

---

# Complete End-to-End Practitioner Case Study

We conclude with a complete end-to-end script illustrating the recommended `mapAI` workflow on historical spatial data:

```r
library(mapAI)
library(sf)
library(terra)

# Step 1: Ingest Ground Control Points with known CRS
data(basel_frickthal)
cps <- basel_frickthal$gcp

# Step 2: Integrated Model Training with Spatial Block Cross-Validation
# Using spatial blocks to prevent spatial autocorrelation data leakage
pai_mod <- train_pai_model(
  gcp_data = cps,
  method = "gam_biv",
  cv = list(validation_type = "spatial_block", k_folds = 5)
)

cat("Spatial Block CV 2D RMSE:", round(pai_mod$cv_rmse_2d, 2), "meters\n")

# Step 3: Diagnostic Distortion and Inversion Screening
dist_diag <- analyze_distortion(pai_model = pai_mod)
if (any(dist_diag$is_inverted)) {
  warning("Topological fold-overs detected! Consider increasing regularization.")
}

# Step 4: Vector Feature Transformation with Automated Topology Repair
parcels_transformed <- transform_map(
  map = historical_parcels,
  model = pai_mod,
  repair_topology = TRUE
)

# Step 5: High-Performance In-Memory Raster Rectification
historical_sheet_rectified <- apply_pai_raster(
  raster = scanned_map_sheet,
  model = pai_mod,
  mesh_size = 50,
  max_iter = 15,
  tol = 1e-4
)
```

---

# References

1. **Bookstein, F. L. (1989)**. Principal warps: Thin-plate splines and the decomposition of deformations. *IEEE Transactions on Pattern Analysis and Machine Intelligence*, 11(6), 567–585.
2. **Golub, G. H., & Van Loan, C. F. (2013)**. *Matrix Computations* (4th ed.). Johns Hopkins University Press.
3. **Roberts, D. R., Bahn, V., Ciuti, S., Boyce, M. S., Elith, J., Guillera-Arroita, G., ... & Dormann, C. F. (2017)**. Cross-validation strategies for data with temporal, spatial, hierarchical or phylogenetic structure. *Ecography*, 40(8), 913–929.
4. **Tissot, A. (1881)**. *Mémoire sur la représentation des surfaces et les projections des cartes géographiques*. Gauthier-Villars.
5. **Tobler, W. R. (1970)**. A computer movie simulating urban growth in the Detroit region. *Economic Geography*, 46(sup1), 234–240.
6. **Valente, R., Vantas, K., & Carrera-Hernández, J. (2021)**. Positional accuracy assessment and improvement of historical maps using non-linear spline models. *International Journal of Geographical Information Science*.
7. **Wood, S. N. (2003)**. Thin plate regression splines. *Journal of the Royal Statistical Society: Series B (Statistical Methodology)*, 65(1), 95–114.
