---
title: "Advanced Analysis with Basel and Frickthal Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{swiss_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette demonstrates a  workflow for analyzing and correcting a historical map with complex, non-linear distortions, focusing on a real-world scenario using the `basel_frickthal` dataset (W. Haas's 1798 map of the Basel and Frickthal region).

The key feature of this dataset is that the gross, linear distortions have already been removed with a Helmert transformation. The remaining errors are subtle, non-linear, and spatially variable. Our goal is to assess different validation techniques, train a robust model for these non-linear residuals, and conduct an in-depth analysis of the spatial distortion.

### Setup

First, we load all the necessary libraries for our workflow.

```{r setup}
# Load the necessary libraries
library(mapAI)
library(sf)
library(dplyr)
library(ggplot2)
library(knitr)
```

### Initial Data Analysis

We begin by loading the `basel_frickthal` dataset and taking its `gcp` element (assigned below to `swiss_cps`). It is a `gcp` object -- a data frame, not an `sf` object -- holding 343 control points whose `source` coordinates have been pre-aligned, so we can focus on the residual `dx` and `dy` errors.

```{r load_data}
# Load the built-in Swiss control points dataset
data(basel_frickthal)
swiss_cps <- basel_frickthal$gcp

# Print a sumary of the control points
summary(swiss_cps)

# Visualize the residual displacement vectors
plot(swiss_cps, title = "Displacement Vectors") +
  labs(title = "", subtitle = "",
       x = "x (m)",
       y = "y (m)")
```

The plot shows clear spatial patterns in the remaining errors, which a simple linear model could not fix. This confirms the need for a non-linear approach.



### Error Assessment and Spatial Validation Schemes

In spatial statistics, standard random cross-validation often suffers from **optimism bias**: close spatial proximity between training and testing observations allows spatial autocorrelation to leak geographic information, artificially depressing estimated prediction errors (Roberts et al., 2017).

To evaluate model generalizability rigorously, `mapAI` implements six distinct validation strategies:

1. **Standard Random CV (`"random"`):** Traditional $k$-fold cross-validation with randomly partitioned points.
2. **Spatial Clustered CV (`"spatial"`):** $k$-means clustering of coordinates to ensure folds form distinct geographic clusters.
3. **Spatial Block CV (`"spatial_block"`):** Rectangular grid partitioning of the geographic bounding box into independent spatial blocks.
4. **Spatial Buffered Block CV (`"spatial_buffered"`):** Imposes an exclusion dead-zone buffer around each test block, discarding nearby training points to reduce spatial autocorrelation data leakage. The buffer removes leakage only if it exceeds the range of residual spatial autocorrelation.
5. **Single Holdout (`"probability"`):** One train/validation split by simple random sampling. Not cross-validation: no replication and no dispersion estimate.
6. **Stratified Sampling (`"stratified"`):** $k$-fold cross-validation binned across distortion magnitude quantiles.

```{r assess_model, message=FALSE}
# Evaluate all six validation strategies using GAM
cv_rnd <- cv_pai_model(swiss_cps, method = "gam_biv", validation_type = "random", k_folds = 5, seed = 42)
cv_spt <- cv_pai_model(swiss_cps, method = "gam_biv", validation_type = "spatial", k_folds = 5, seed = 42)
cv_blk <- cv_pai_model(swiss_cps, method = "gam_biv", validation_type = "spatial_block", k_folds = 5, seed = 42)
cv_buf <- cv_pai_model(swiss_cps, method = "gam_biv", validation_type = "spatial_buffered", k_folds = 5, seed = 42)
cv_prb <- cv_pai_model(swiss_cps, method = "gam_biv", validation_type = "probability", train_split_ratio = 0.8, seed = 42)
cv_str <- cv_pai_model(swiss_cps, method = "gam_biv", validation_type = "stratified", k_folds = 5, seed = 42)

# Combine results into a comprehensive comparison table
validation_df <- rbind(
  cv_rnd$summary, cv_spt$summary, cv_blk$summary,
  cv_buf$summary, cv_prb$summary, cv_str$summary
) %>%
  select(Model_Method = Method, Mean_2D_RMSE_m = Mean_RMSE_2D, Std_Deviation_of_RMSE_m = SD_RMSE_2D) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

validation_df$Strategy <- c(
  "Standard Random 5-fold CV",
  "Spatial Clustered 5-fold CV",
  "Spatial Block 5-fold CV",
  "Spatial Buffered Block 5-fold CV",
  "Single 80/20 Holdout (not CV)",
  "Stratified 5-fold CV"
)

kable(validation_df[, c("Strategy", "Model_Method", "Mean_2D_RMSE_m", "Std_Deviation_of_RMSE_m")],
      caption = "Table: Predictive performance comparison across six validation strategies.")
```

The spatial schemes report higher errors than random cross-validation because each fold is evaluated on held-out geographic regions rather than on points surrounded by their own neighbours. The spread across schemes is itself the result: it measures how sensitive the accuracy estimate is to the partitioning, and no single row is the one "true" out-of-sample number. Report the scheme matching the intended use, state `block_size` and `buffer_dist`, and bear in mind that these schemes reduce autocorrelation optimism without removing it. The 80/20 row is a single holdout with no replication, so its `NA` dispersion is not comparable with the k-fold rows.

### Training with Integrated Spatial Cross-Validation

Rather than running validation as an external detached step, `train_pai_model` allows cross-validation to be integrated directly into model fitting via the `cv` argument. This ensures that out-of-sample validation metrics (`cv_rmse_2d`) are permanently packaged with the fitted model:

```{r trainmodel, message=FALSE}
# Train final GAM model with integrated 5-fold spatial block cross-validation
final_gam_model <- train_pai_model(
  gcp_data = swiss_cps,
  method = "gam_biv",
  cv = list(validation_type = "spatial_block", k_folds = 5, seed = 42)
)

# Inspect model summary showing both in-sample training and out-of-sample CV metrics
print(final_gam_model)

# Use the regular line grid to serve as our reference map
source_grid <- basel_frickthal$grid 

# Apply the trained model to correct the grid with topology validation
gam_corrected_grid <- transform_map(final_gam_model, source_grid, repair_topology = TRUE)

# Prepare data for plotting
source_grid$status <- "Original (Distorted)"
gam_corrected_grid$status <- "GAM"
comparison_data  <- rbind(source_grid[, "status"], gam_corrected_grid[, "status"])

# Create the comparison plot
ggplot(comparison_data) +
  geom_sf(aes(color = status, linetype = status), fill = NA, linewidth = 0.7) +
  scale_color_manual(name = "Map Status", values = c("Original (Distorted)" = "grey50", "GAM" = "#e41a1c")) +
  geom_point(data = swiss_cps, aes(source_x, source_y), size = 0.2) +
  scale_linetype_manual(name = "Map Status", values = c("Original (Distorted)" = "dashed", "GAM" = "solid")) +
  labs(x = "x (m)", y = "y (m)") +
  coord_sf(datum = sf::st_crs(21781)) +
  ggplot2::theme_minimal()
```

The plot shows how the regular grid (dashed grey) is warped by the model into the corrected grid (solid red). This transformation is precisely what's needed to align the distorted map features with their true locations.



### Residual Analysis

After applying the correction, we should examine the remaining errors, or residuals. This helps us confirm that our model has captured the systematic distortion patterns and that the remaining errors are small and random.

```{r residuals}
# In-sample residuals: the model evaluated on its own training points.
residuals(final_gam_model)

# Out-of-fold residuals: the honest picture of predictive error. Available
# because the model was trained with `cv`.
residuals(final_gam_model, out_of_sample = TRUE)
```

The two plots answer different questions. The first shows what the model failed
to absorb on the points it was fitted to; for an exact interpolator such as
`tin_linear` it would be empty by construction. The second shows the error at
points the model never saw, and is the one to judge predictive accuracy by.

The residual vectors are much smaller and show no obvious spatial pattern, indicating a successful model fit.



### Advanced Distortion Analysis

To quantify the nature of the distortion, we evaluate `analyze_distortion()` at the 343 control point locations. (Pass a `terra::SpatRaster` instead to obtain a continuous, gridded distortion surface.) We will visualize three key metrics:

  * **Area Scale**: Shows how much areas are stretched or compressed. A value of 1 means no distortion.
  * **Maximum Angular Distortion**: Measures the maximum change in angle at any point.
  * **Airy-Kavrayskiy Measure**: A combined measure of both areal and angular distortion.


```{r distortions}
# Evaluate the distortion metrics at the control point locations
distortion_results <- analyze_distortion(final_gam_model, swiss_cps)

# Plot 1: Area scale distortion
plot(distortion_results, "area_scale", diverging = TRUE) +
  labs(title = "Area Scale Distortion", subtitle = "", x = "X (m)", y = "Y (m)", fill = expression(a %*% b))

# Plot 2: Maximum angular distortion
plot(distortion_results, "max_angular_distortion", palette = "magma") +
  labs(title = "Maximum Angular Distortion", subtitle = "", x = "X (m)", y = "Y (m)", fill = "2Ω (rad)")

# Plot 3: Airy-Kavrayskiy distortion measure
plot(distortion_results, "airy_kavrayskiy") +
  labs(title = "Airy-Kavrayskiy Combined Distortion", subtitle = "", x = "x (m)", y = "y (m)", fill = expression(E["AK"]))
```

We can also summarize these distortion metrics numerically.

```{r summary}
# Calculate and print summary statistics for the distortion metrics
print(kable(summary(distortion_results), caption = "Table: Summary of distortion metrics from GAM model analysis."))
```


### Visualizing Distortion with Indicatrices

A classic way to visualize map distortion is with **Tissot's Indicatrix**, which shows how an infinitesimal circle on the original surface is deformed into an ellipse. We can plot these at our control point locations to see the local distortion characteristics.

```{r indicatrices}

# Plot the indicatrices (ellipses) at each point
indicatrices(distortion_results,  scale_factor = 1000) +
  labs(
    title = "Distortion Indicatrices at Control Points",
    subtitle = "",
    x = "x (m)",
    y = "y (m)"
  )
```

The size and shape of the ellipses clearly show the spatial variation in both the magnitude and direction of the distortion that our model has successfully captured.

### Checking for Topological Fold-Overs

An essential check when applying non-linear spline models is ensuring that the deformation has not introduced local topological fold-overs or singular collapses. We verify this by inspecting the signed Jacobian determinant $\det(\mathbf{J})$:

```{r check_inversion}
# Verify that all points maintain a positive Jacobian determinant
n_inverted <- sum(distortion_results$is_inverted, na.rm = TRUE)
cat(sprintf("Topological fold-overs detected: %d (min det(J) = %.4f)\n",
            n_inverted, min(distortion_results$det_J, na.rm = TRUE)))
```

Two caveats on how far this screening reaches. First, $\det(\mathbf{J})$ was evaluated **at the 343 control point locations only**, not continuously: a fold-over between control points would go undetected. Pass a `SpatRaster` to `analyze_distortion()` to screen the domain on a grid.

Second, even a uniformly positive Jacobian determinant establishes only that the map is a **local** diffeomorphism. It does not imply global injectivity — a map can have $\det(\mathbf{J}) > 0$ everywhere and still fold onto itself globally (the standard counterexample being $z \mapsto e^z$ in the complex plane). So a positive determinant at every sampled point is reassuring evidence against local fold-over, not a guarantee of global invertibility.

---

### References

- Roberts, D. R., Bahn, V., Ciuti, S., Boyce, M. S., Elith, J., Guillera-Arroita, G., Hauenstein, S., Lahoz-Monfort, J. J., Schröder, B., Thuiller, W., Warton, D. I., Wintle, B. A., Hartig, F., & Dormann, C. F. (2017). Cross-validation strategies for data with spatial, temporal, or phylogenetic dependence. *Ecography*, 40(8), 913-929. https://doi.org/10.1111/ecog.02881
- Snyder, J. P. (1987). *Map Projections: A Working Manual*. U.S. Geological Survey Professional Paper 1395.
- Tissot, A. (1881). *Mémoire sur la représentation des surfaces et les projections des cartes géographiques*. Gauthier-Villars.
- Valavi, R., Elith, J., Lahoz-Monfort, J. J., & Guillera-Arroita, G. (2019). blockCV: An R package for generating spatially or environmentally separated folds for k-fold cross-validation of species distribution models. *Methods in Ecology and Evolution*, 10(2), 225-232.
- Vantas, K., & Mirkopoulou, E. (2025). *mapAI: An R Package for Positional Accuracy Improvement of Vector Maps*.
- Wood, S. N. (2017). *Generalized Additive Models: An Introduction with R* (2nd ed.). Chapman & Hall/CRC.
